---
title: "Chapter 15 - Functions"
author: "Vebash Naidoo"
date: "06/11/2020"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE, message=FALSE,
                      comment = "#>", 
                      collapse = TRUE)

options(scipen=10000)
library(tidyverse)
library(flair)
library(magrittr)
```

# Functions

```{r rep1, include = FALSE}
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
  (max(df$b, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) / 
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))
```

```{r, echo=FALSE}
decorate("rep1") %>% 
  flair('max(df$b, na.rm = TRUE) - min(df$a, na.rm = TRUE)',
        background = "#E3242B", 
        color = "#FFE4E1")
```

Note the mistake in the copy and paste - we're using `df$a` for the calculation
for `df$b`!


Let's isolate the main functionality that we are __repeating__.

```{r}
(df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
```

That's all good and well, but to write a function you should generalise the
functionality as a test first.

```{r}
x <- df$a

(x - min(x, na.rm = TRUE)) /
  (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
```

Even better, we're using the `range` when we use `max` and `min`. So let's re-write using that.

```{r}
rng <- range(x, na.rm=TRUE)
(x - rng[1]) /
  (rng[2] - rng[1])

```

Let's pull it all together in a <span style="color: #008080;background-color:#9FDDBA">function</span>.

```{r fun1, include=FALSE}
rescale01 <- function(x) {
  # rescales a vector to lie between 0 and 1
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) /(rng[2] - rng[1])
}

rescale01(c(0,5,10))
```

```{r, echo=FALSE}
decorate("fun1") %>% 
  flair("function", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("{", background = "#cfdad4", 
        color = "#6d6875") %>% 
  flair("}", background = "#cfdad4", 
        color = "#6d6875")
```

Key steps:

1.  Pick a 
<span style="color: #008080;background-color:#9FDDBA">__name__</span> for the function. 

1.  List the inputs, or __arguments__, to the function inside 
<span style="color: #008080;background-color:#9FDDBA">`function`</span>.
    Example: 
<span style="color: #008080;background-color:#9FDDBA">`function(x, y, z)`</span> .

1.  Place your code in __body__ of the function, a 
<span style="color: #008080;background-color:#9FDDBA">`{`</span>  block that immediately follows `function(...)`.

1. Make it work with a small input.

1. Check the function with a range of inputs.

```{r}
rescale01(c(-10,0,10))

rescale01(c(1,2,3,NA,5))
```

Simplifying the original calcs:

```{r}
df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

Now changes can be made in a single place.

```{r}
x <- c(1:10, Inf)
rescale01(x)
```

```{r}
rescale01 <- function(x){
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) /(rng[2] - rng[1])
}

rescale01(x)
```

### Intro - Exercises

1.  Why is `TRUE` not a parameter to `rescale01()`? What would happen if
    `x` contained a single missing value, and `na.rm` was `FALSE`?

    ```{r}
    rescale01_miss <- function(x){
      rng <- range(x, na.rm = FALSE)
      (x - rng[1]) / (rng[2] - rng[1])
    }
    rescale01_miss(c(1:5, NA, 10))
    
    rescale01_miss <- function(x){
      rng <- range(x, na.rm = FALSE, finite = TRUE)
      (x - rng[1]) / (rng[2] - rng[1])
    }
    rescale01_miss(c(1:5, NA, 10))
    
    ```
    
    If any value is NA the range function returns NA as min and max of
    range. If `finite = TRUE` is set only the NA value returns NA.
    
    According to the help page: "If finite is TRUE, the minimum and maximum 
    of all finite values is computed, i.e., `finite = TRUE` includes 
    `na.rm = TRUE`."
    

1.  In the second variant of `rescale01()`, infinite values are left
    unchanged. Rewrite `rescale01()` so that `-Inf` is mapped to 0, and 
    `Inf` is mapped to 1.
    
    ```{r}
    rescale01_inf <- function(x){
      rng <- range(x, na.rm = TRUE, finite = TRUE)
      x[x == -Inf] <- 0
      x[x == Inf] <- 1
      (x - rng[1]) / (rng[2] - rng[1])
    }
    
    rescale01_inf(c(1, Inf, 0, -Inf))
    ```
    

1.  Practice turning the following code snippets into functions. Think about 
    what each function does. What would you call it? How many arguments does it
    need? Can you rewrite it to be more expressive or less duplicative?

    ```{r, eval = FALSE}
    mean(is.na(x))
    
    x / sum(x, na.rm = TRUE)
    
    sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)
    ```
    
    `mean(is.na(x))` finds what percentage of the data is NA.
    
    ```{r}
    x <- c(NA, 1:10, NA, 5)
    mean(is.na(x))
    
    mean_na <- function(x){
      sum(x)/length(x)
    }
    mean_na(is.na(x))
    ```
    
    `x / sum(x, na.rm = TRUE)` converts each x to a percentage it contributes 
    towards 100%, not considering the NAs.
    
    ```{r}
    contrib_to_one <- function(x) {
      x / sum(x, na.rm = TRUE)
    }

    contrib_to_one(x)
    ```
    
    `sd(x, na.rm = TRUE) / mean(x, na.rm = TRUE)` finds how the deviation
    as a proportion of the mean. `sd` means standard deviation and 
    is a measure of the variability in your data. The `mean` is the average
    of your data.
    
    ```{r}
    deviation <- function(x, na.rm = TRUE) {
      sd(x, na.rm = na.rm) / mean(x, na.rm = na.rm)
    }
    
    deviation(x)
    ```
    
    

1.  Write your own functions to compute the variance and skewness of a numeric vector.
    Variance is defined as
    $$
    \mathrm{Var}(x) = \frac{1}{n - 1} \sum_{i=1}^n (x_i - \bar{x}) ^2 \text{,}
    $$
    where $\bar{x} = (\sum_i^n x_i) / n$ is the sample mean.
    Skewness is defined as
    $$
    \mathrm{Skew}(x) = \frac{\frac{1}{n-2}\left(\sum_{i=1}^n(x_i - \bar x)^3\right)}{\mathrm{Var}(x)^{3/2}} \text{.}
    $$
    
    ```{r}
    
    variance <- function(x){
      # find non na length of vector x
      n <- sum(!is.na(x))
      avg <- mean(x, na.rm = TRUE)
      (1/(n-1))*(sum((x - avg)^2, na.rm=TRUE))
    }
    
    x <- c(1:10)
    variance(x)
    var(x, na.rm = TRUE)
    
    y <- c(1:5, 1:5)
    variance(y)
    var(y)
    
    z <- c(1:10, NA)
    variance(z)
    var(z, na.rm = TRUE)
    ```
    
    ```{r}
    skew <- function(x){
        n <- sum(!is.na(x))
        avg <- mean(x, na.rm = TRUE) 
        summation <- sum((x - avg)^3, na.rm = TRUE)
        variance_val <- (variance(x))^(3/2)
        ((1/(n-2))*summation)/variance_val
    }
    
    a <- c(1:10, 1000)
    skew(a)
    ```

1.  Write `both_na()`, a function that takes two vectors of the same length 
    and returns the number of positions that have an `NA` in both vectors.
    
    ```{r}
    both_na <- function(x, y){
      # find where x is NA - is.na(x) - returns a bunch of TRUE, FALSE
      # find where y is NA - is.na(y) - returns a bunch of TRUE, FALSE
      # where are they both NA - &
      # sum where they are both NA 
      # e.g. (T, F, T) & (T, F, F) == 1; only T in same place once
      sum(is.na(x) & is.na(y))
    }
    
    x <- c(1:10, NA, 12, NA)
    y <- c(1:9, 10, NA, 12, 13)
    both_na(x,y)
    both_na(c(NA, 2, 4, 6, NA, 10, 12),
            c(NA, 1, 3, 5, NA, 9, 11))
    ```
    

1.  What do the following functions do? Why are they useful even though they
    are so short?
    
    - `is_directory()` tells you whether a given path is a directory.
    - `is_readable()` tells you whether you have permission to read
    a file.
    
      ```{r}
      is_directory <- function(x) file.info(x)$isdir
      is_directory("C:/Personal")
      is_directory("C:/Work/Learning/how-containers-work.pdf")
      is_readable <- function(x) file.access(x, 4) == 0
      is_readable("C:/Personal")
      is_readable("C:/Work/Learning/how-containers-work.pdf")
      ```

1.  Read the [complete lyrics](https://en.wikipedia.org/wiki/Little_Bunny_Foo_Foo) 
    to "Little Bunny Foo Foo". There's a lot of duplication in this song. 
    Extend the initial piping example to recreate the complete song, and use 
    functions to reduce the duplication.
    
    <pre>
    Repeat (x3)
    Little Bunny Foo Foo,
    Hopping through the forest,
    Scooping up the field mice,
    And bopping them on the head.

    (Spoken)
    Down came the Good Fairy, and she said,

    "Little Bunny Foo Foo,
    I don't want to see you,
    Scooping up the field mice
    And bopping them on the head."

    (Spoken)
    "I'll give you three (two / one) chance(s),
    And if you don't behave,
    I'm gonna turn you into a goon!"
    END Repeat

    The next day... or That evening... or Later that night...
    "I gave you three chances,
    And you didn't behave,
    And now I'm gonna turn you into a goon. POOF!"
    
    "And the moral of the story is: Hare today, goon tomorrow."
    </pre>
    
    ```{r}
    little_bunny <- function(bunny = ""){
      str_glue("{bunny}Little Bunny Foo-Foo,\n\n")
    }
    
    hop <- function(bunny, where = "forest"){
      str_glue("{bunny}Hopping through the {where},\n\n")
    }
    
    scoop <- function(bunny, what = "field mice"){
      str_glue("{bunny}Scooping up the {what},\n\n")
    }
    
    bop <- function(bunny, where = "head"){
      str_glue("{bunny}And bopping them on the {where}.\n\n\n")
    }
    
    chances <- function(bunny, num = "three"){
      if(num == "one") {
        str_glue("{bunny}I'll give you {num} chance,\n\n")
      }
      else {
        str_glue("{bunny}I'll give you {num} chances,\n\n")
      }
    }
    
    static_line1 <- function(bunny){
      str_glue("{bunny}I don't want to see you,\n\n")
    }
    
    static_line2 <- function(bunny){
      str_glue("{bunny}And if you don't behave,\nI'm gonna turn you into a goon!\n\n\n")
    }
    
    static_line3 <- function(bunny){
      str_glue("{bunny}Down came the Good Fairy, and she said,\n\n")
    }
    
    
    foo_foo <- little_bunny()
    foo_foo %>% 
      hop(where = "forest") %>% 
      scoop(what = "field mice") %>% 
      bop(where = "head") %>% 
      static_line3() %>% 
      little_bunny() %>% 
      static_line1() %>% 
      scoop(what = "field mice") %>% 
      bop(where = "head") %>%
      chances(num = "three") %>% 
      static_line2() %>%
      little_bunny() %>% 
      hop(where = "forest") %>% 
      scoop(what = "field mice") %>% 
      bop(where = "head") %>% 
      static_line3()%>% 
      little_bunny() %>% 
      static_line1()%>% 
      scoop(what = "field mice") %>% 
      bop(where = "head") %>%
      chances(num = "two") %>% 
      static_line2() %>% 
      little_bunny() %>% 
      hop(where = "forest") %>% 
      scoop(what = "field mice") %>% 
      bop(where = "head") %>% 
      static_line3()%>% 
      little_bunny() %>% 
      static_line1() %>% 
      scoop(what = "field mice") %>% 
      bop(where = "head") %>%
      chances(num = "one") %>% 
      static_line2() %>%
      little_bunny() %>% 
      hop(where = "forest") %>% 
      scoop(what = "field mice") %>% 
      bop(where = "head") %>% 
      static_line3() %>%   
      str_glue("I gave you three chances,\nAnd you didn't behave,\nAnd now I'm gonna turn you into a goon. POOF!")
    ```


### Human Readable Functions - Exercises

1.  Read the source code for each of the following three functions, puzzle out
    what they do, and then brainstorm better names.
    
    ```{r}
    f1 <- function(string, prefix) {
      substr(string, 1, nchar(prefix)) == prefix
    }
    f2 <- function(x) {
      if (length(x) <= 1) return(NULL)
      x[-length(x)]
    }
    f3 <- function(x, y) {
      rep(y, length.out = length(x))
    }
    ```
    
    The first function `f1` checks if the substring of string
    matches the prefix given.
    
    ```{r}
    match_prefix <- function(string, prefix) {
      substr(string, 1, nchar(prefix)) == prefix
    }
    match_prefix("===The quick brown fox", "===")
    match_prefix("```The quick brown fox", "===")
    match_prefix(c("the man in the bowler hat",
                   "that quick brown fox",
                   "the one and only",
                   "that crafy ol' fox"),
                 "the")
    ```
    
    The function removes the last item of vector x.
    
    ```{r}
    remove_last <- function(x) {
      if (length(x) <= 1) return(NULL)
      x[-length(x)]
    }
    remove_last(c(1:5, 2, 30, 45, NA))
    remove_last(c(NA, 2, 30, 45, 1:5))
    ```
    
    The function repeats the y vector as many times as the length
    of vector x.
    
    ```{r}
    rep_len <- function(x, y) {
      rep(y, length.out = length(x))
    }
    rep_len(c(1:10), 1)
    rep_len(c("a", "b", "abc", "abcd", "efghi"),
            c("a", "b", "abc"))
    ```
    
1.  Take a function that you've written recently and spend 5 minutes 
    brainstorming a better name for it and its arguments.

1.  Compare and contrast `rnorm()` and `MASS::mvrnorm()`. How could you make
    them more consistent? 
    
    |rnorm()            |    MASS::mvrnorm() |
    |:------------------|:-------------------|
    |univariate normal distribution  |multivariate normal distribution   |
    |n, mean, sd | n, mu, Sigma |
    
    You could make mean / mu and sd / Sigma consistent by choosing one of these
    names. I would also make all names lower case so I would change Sigma to
    sigma.
    

1.  Make a case for why `norm_r()`, `norm_d()` etc would be better than
    `rnorm()`, `dnorm()`. Make a case for the opposite.
    
    - For `norm_r()`, `norm_d()` the fact that you can type `norm` and the
    autocomplete will give you a list of all `norm_` functions is
    a win.
    
    - For `rnorm()`, `dnorm()` I would say it is more natural to say
    *random normal distribution* / *density normal distribution* than 
    "normal random distribution" / "normal density distribution".
    
### Conditional execution

An `if` statement allows conditional execution of code.

```{r, eval = FALSE}
if (condition) {
  # code executed when condition is TRUE
} else {
  # code executed when condition is FALSE
}
```

- To get help on `if` you need to surround it in backticks: `` ?`if` ``. 
- The `condition` must evaluate to either `TRUE` or `FALSE`. You will get a warning / error if it is a vector OR if it's an `NA`.
- Use `||` (**or**) and `&&` (__and__) to combine multiple logical expressions. These operators are "short-circuiting"
- __NEVER__ use `|` or `&` in an `if` statement, since these are vectorised
operations.
- `==` is also vectorised, which means you may get more than one output!
  
  * Check length is 1 already.
  * Collapse with `any()` or `all()`.
  * Use the non-vectorised `identical()`, __BUT__ this is really strict and
  can result in unexpected output given how computers store numbers etc.
  Use `dplyr::near()` for comparisons.
  * Use `is.na()` for NA checks.

```{r, error = TRUE}
if (c(TRUE, FALSE)) {}
if (NA) {}
```

```{r}
identical(0L, 0)
```

```{r}
x <- sqrt(2) ^ 2
x
x == 2
x - 2
```

Multiple if statements are also allowed.

```{r, eval = FALSE}
if (this) {
  # do that
} else if (that) {
  # do something else
} else {
  # 
}
```

Another useful technique is the `switch()` function for many `if ... else` constructs.

```{r, warning = TRUE, error = TRUE}
perform_calc <- function(x, y, op) {
  switch(op,
    plus = x + y,
    minus = x - y,
    times = x * y,
    divide = x / y,
    stop("Unknown op!")
  )
}
perform_calc(c(1,2,3), c(4,5,6), "plus")
perform_calc(c(1,2,3), c(4,5), "minus")
perform_calc(c(1,2,3), c(4,5,6), "power")
```

### Conditions - Exercises

1.  What's the difference between `if` and `ifelse()`? Carefully read the help
    and construct three examples that illustrate the key differences.
    
    `if` tests just one item, so it has to be used in a loop or
    with `any()` / `all()` if you want to ensure all in a vector
    meet a condition or any meet a condition.
    
    `ifelse` checks each item in the vector.
    
    If you read the help [?`ifelse`] it says use `if`/`else` construct for 
    simple yes/no answers, and use `ifelse` for over a set of values.
    
    ```{r ifelse1, warning=TRUE, include=FALSE}
    x <- c(1:5, 20, 25, 30)
    
    if(length(x) > 10){
      print(length(x))
      TRUE
    } else {
      print(length(x))
      FALSE
    }
    
    # You can use ifelse but kinda silly to use in this case
    ifelse(length(x)>10, length(x), "smaller than 10 items in vector")
    
    # here is a better use
    ifelse(x > 5, TRUE, FALSE)
    
    # DOES NOT work for if --- notice the warning that only
    # 1st is used
    if (x > 5) TRUE else FALSE
    ```
    
    ```{r, echo=FALSE}
    decorate("ifelse1") %>% 
      flair('DOES NOT work',
            background = "#cfdad4", 
            color = "#6d6875")
    ```
    
1.  Write a greeting function that says "good morning", "good afternoon",
    or "good evening", depending on the time of day. (Hint: use a time
    argument that defaults to `lubridate::now()`. That will make it 
    easier to test your function.)
    
    ```{r}
    greeting <- function(){
      time <- lubridate::now()
      hr <- lubridate::hour(time)
      if(hr >= 0 && hr < 12){
        "good morning"
      } else if (hr >= 12 && hr < 18){
        "good afternoon"
      } else {
        "good evening"
      }
    }
    greeting()
    ```
    
1.  Implement a `fizzbuzz` function. It takes a single number as input. If
    the number is divisible by three, it returns "fizz". If it's divisible by
    five it returns "buzz". If it's divisible by three and five, it returns
    "fizzbuzz". Otherwise, it returns the number. Make sure you first write 
    working code before you create the function.
    
    ```{r}
    fizzbuzz <- function(x){
      if((x %% 3 == 0) && (x %% 5 == 0)){
        "fizzbuzz"
      } else if (x %% 3 == 0) {
        "fizz"
      } else if (x %% 5 == 0){
        "buzz"
      } else {
        x
      }
    }
    fizzbuzz(5)
    fizzbuzz(9)
    fizzbuzz(30)
    fizzbuzz(10)
    ```
    
1.  How could you use `cut()` to simplify this set of nested if-else statements?

    ```{r, warning = TRUE}
    temp <- 31
    if (temp <= 0) {
      "freezing"
    } else if (temp <= 10) {
      "cold"
    } else if (temp <= 20) {
      "cool"
    } else if (temp <= 30) {
      "warm"
    } else {
      "hot"
    }
    
    cut(temp, breaks = 10*(-3:10), 
        labels = c(rep("freezing", 3),
                   "cold",
                   "cool",
                   "warm",
                   rep("hot", 7)))
    
    table(cut(temp, breaks = 10*(-3:10), 
        labels = c(rep("freezing", 3),
                   "cold",
                   "cool",
                   "warm",
                   rep("hot", 7))))
    ```
    
    How would you change the call to `cut()` if I'd used `<` instead of `<=`?
    What is the other chief advantage of `cut()` for this problem? (Hint:
    what happens if you have many values in `temp`?)
    
    You would use `right = FALSE`. `cut` allows vectors of values whereas
    `if else` constructs do not.
    
    __NOTE__: If you look at [jrnold's solutions](https://jrnold.github.io/r4ds-exercise-solutions/functions.html) 
    you note that he uses `-Inf` and `Inf` for the breaks. This is much better
    than my hacky solution above that contains boundaries at the bottom and
    top.

1.  What happens if you use `switch()` with numeric values?

    Read more [here](https://www.datamentor.io/r-programming/switch-function/).
    
    `switch` evaluates the `EXPR` against the list item number. Below we have 3
    and this evaluates to list item == 3, hence 6.
    
    ```{r}
    x <-  3
    switch(x, 
      "2",
      "4",
      "6",
      "8"
    )
    ```


1.  What does this `switch()` call do? What happens if `x` is "e"?

    If x is a or b, it returns "ab"; if x is c or d it returns cd. If x
    is e, no match is found and nothing is returned.
    
    The docs say to have an unnamed value as default after the named
    values for no match ones. This can be done as per the last way - note the
    <span style="color: #6d6875;background-color:#cfdad4">test</span>.

    ```{r, eval = FALSE}
    x <-  "c"
    switch(x, 
      a = ,
      b = "ab",
      c = ,
      d = "cd"
    )
    
    x <- "e"
    switch(x, 
      a = ,
      b = "ab",
      c = ,
      d = "cd"
    )
    
    x <- "e"
    switch(x, 
      a = ,
      b = "ab",
      c = ,
      d = "cd"
      "test"
    )
    ```
    
    Experiment, then carefully read the documentation. 

#### Default Arguments

The `data` to apply a function to should come first, followed by the 
arguments. Arguments should have default values as far as possible.

```{r def1}
# Compute confidence interval around mean using normal approximation
mean_ci <- function(x, conf = 0.95) {
  se <- sd(x) / sqrt(length(x))
  alpha <- 1 - conf
  mean(x) + se * qnorm(c(alpha / 2, 1 - alpha / 2))
}
x <- runif(100)
mean_ci(x)
mean_ci(x, conf = 0.99)
```

```{r, echo=FALSE}
decorate("def1") %>% 
  flair("conf = 0.95", background = "#9FDDBA", 
        color = "#008080")
```

 It's good practice to check important preconditions, and throw an error (with `stop()`), if they are not met.

```{r, error = TRUE}
wt_mean <- function(x, w) {
  if (length(x) != length(w)) {
    stop("`x` and `w` must be the same length", call. = FALSE)
  }
  sum(w * x) / sum(w)
}
wt_mean(1:6, 1:3)
```

A useful function is the built-in `stopifnot()` which checks that each 
argument is `TRUE`, and produces a generic error message if not.

```{r, error = TRUE}
wt_mean <- function(x, w, na.rm = FALSE) {
  # I want na.rm to be logical: is.logical(na.rm)
  # I want length na.rm to be 1: length(na.rm) == 1
  # If it is NOT that STOP!
  stopifnot(is.logical(na.rm), length(na.rm) == 1)
  stopifnot(length(x) == length(w))
  
  if (na.rm) {
    miss <- is.na(x) | is.na(w)
    x <- x[!miss]
    w <- w[!miss]
  }
  sum(w * x) / sum(w)
}
wt_mean(1:6, 6:1, na.rm = "foo")
```

__NOTE__: When using `stopifnot()` you assert what should be true rather than checking for what might be wrong.

#### Dot-dot-dot (...)

Many functions in R take an arbitrary number of inputs by using a special argument: `...` (pronounced dot-dot-dot). You can forward `...` to other functions.

```{r}
sum(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
stringr::str_c("a", "b", "c", "d", "e", "f")
```

```{r}
commas <- function(...) stringr::str_c(..., collapse = ", ")
commas(letters[1:10])
rule <- function(..., pad = "-") {
  title <- paste0(...)
  # getOption("width") tells you how many chars can be printed
  # nchar(title) tells you how long title is
  width <- getOption("width") - nchar(title) - 5
  cat(title, " ", stringr::str_dup(pad, width), "\n", sep = "")
}
rule("Important output")
```

```{r wrong, include = FALSE}
x <- c(1, 2)
sum(x, na.mr = TRUE)
```

```{r, echo=FALSE}
decorate("wrong") %>% 
  flair("na.mr", background = "#ffa578", 
        color = "#FFE4E1")  
```

Notice the result above is incorrect as well, since it sums `1 + 2 + TRUE` = 4,
but it should be 3!


### Func Arguments - Exercises

1.  What does `commas(letters, collapse = "-")` do? Why?

    It throws an error. The reason is that both `letters` and `collapse` are
    passed into commas in the `...` argument. 
    
    ```{r, error = TRUE}
    commas(letters, collapse = "-")
    ```

1.  It'd be nice if you could supply multiple characters to the `pad` argument, 
    e.g. `rule("Title", pad = "-+")`. Why doesn't this currently work? How 
    could you fix it?

    ```{r}
    rule <- function(..., pad = "-") {
      title <- paste0(...)
      # getOption("width") tells you how many chars can be printed
      # nchar(title) tells you how long title is
      width <- getOption("width") - nchar(title) - 5
      cat(title, " ", stringr::str_dup(pad, width), "\n", sep = "")
    }
    rule("Important output", pad = "-+")
    ```    

    The width of the resulting string is too long this way. 
    
    ```{r rule, include = FALSE}
    rule <- function(..., pad = "-") {
      title <- paste0(...)
      # getOption("width") tells you how many chars can be printed
      # nchar(title) tells you how long title is
      width <- getOption("width") - nchar(title) - 5
      cat(title, " ", stringr::str_dup(pad, width/nchar(pad)), "\n", sep = "")
    }
    rule("Important output", pad = "-+")
    ``` 
    
    ```{r, echo=FALSE}
    decorate("rule") %>% 
      flair("nchar(pad)", background = "#9FDDBA", 
            color = "#008080")
    ```
    
1.  What does the `trim` argument to `mean()` do? When might you use it?

    The `trim` argument removes / trims a certain percentage of the observations
    from each side. In the help page if we consider the examples, notice that
    the 0 and 50 are trimmed. So the values are sorted, and then the trimming
    occurs - this makes sense because we don't want to just arbitrarily chop
    data, we want to remove *outliers*.
    
    In the second example we sort 5, 10, 11, ..., 19, 20. Then 5 and 20 trimmed.
    
    ```{r}
    x <- c(0:10, 50)
    (xm <- mean(x))
    c(xm, mean(x, trim = 0.10))
    mean(c(1:10))
    mean(c(1:10)) == mean(x, trim = 0.10)
    
    x <- c(10:20, 5)
    (xm <- mean(x))
    c(xm, mean(x, trim = 0.10))
    mean(c(10:19))
    mean(c(10:19)) == mean(x, trim = 0.10)
    ```

1.  The default value for the `method` argument to `cor()` is 
    `c("pearson", "kendall", "spearman")`. What does that mean? What 
    value is used by default?
    
    The `cor()` function can use any one of those values, by default the first 
    is used which is pearson.
    


#### Writing pipeable functions

If you want to write your own pipeable functions, think about the return value. E.g. for dplyr and tidyr the object type is the data frame. 

There are two basic types of pipeable functions: __transformations__ and __side-effects__. 

- With __transformations__, an object is passed to the function's first argument and a modified object is returned. 
- With __side-effects__, the passed object is not transformed. Instead, the function performs an action on the object, like printing or plotting. These functions "invisibly" return the first argument, so that they can still be used in a pipeline.

```{r}
show_missings <- function(df) {
  n <- sum(is.na(df))
  cat("Missing values: ", n, "\n", sep = "")
  
  invisible(df) # return df invisibly
}
```

```{r}
# df does not get printed but it is there
show_missings(mtcars)
```

It's still there, it's just not printed by default.

```{r}
x <- show_missings(mtcars) 
class(x)
dim(x)
head(x)
```

We can use it in a pipe:

```{r}
library(dplyr)
mtcars %>% 
  show_missings() %>% 
  mutate(mpg = ifelse(mpg < 20, NA, mpg)) %>% 
  show_missings() 
```

### What's in a function?

Want to see what the code for a function is?

Type the function name with no parantheses.

E.g. `lm`, `factorial`

```{r}
lm
```

To figure out what arguments a function takes use `args(func_name)`.

```{r}
args(lm)
args(dplyr::across)
```


