---
title: "Chapter 17 - Iteration with {purrr}"
author: "Vebash Naidoo"
date: "10/11/2020"
output: html_document
---

```{css, echo = FALSE}
.tabset h2 {display: none;}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE, message=FALSE,
                      comment = "#>", 
                      collapse = TRUE)

options(scipen=10000)
library(tidyverse)
library(flair)
library(magrittr)
library(lubridate)
library(werpals) # my own package - sciencificity github
```
# Iteration {#buttons .tabset .tabset-fade .tabset-pills}

__Click on the tab buttons below for each section__

<h2>For loops</h2>
## For loops

We have seen one tool to avoid repeating yourself `functions`.
Another tool for reducing duplication is __iteration__, which helps you when 
you need to do the same thing to multiple inputs: repeating the same operation 
on different columns, or on different datasets. 

Let's say we want to calculate the median for each column in a dataframe,
and we don't want to repeat ourselves.

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

# what we don't want
median(df$a)
# ...
median(df$d)

df %>% 
  summarise(across(.cols = dplyr::everything(),
                median,
                .names = "median_{.col}"))

output <- vector("double", ncol(df)) #1. output
for (i in seq_along(df)){            #2. sequence 
  output[[i]] <- median(df[[i]])     #3. body
}
output
```

A `for loop` has three components:

1.  The __output__: `output <- vector("double", length(x))`. 
    Before you start the loop, you must always allocate sufficient space 
    for the output. This is very important for efficiency. 
    
    A general way of creating an empty vector of given length is the `vector()`
    function. It has two arguments: the type of the vector ("logical", 
    "integer", "double", "character", etc) and the length of the vector. 

1.  The __sequence__: `i in seq_along(df)`. This determines what to loop over:
    each run of the for loop will assign `i` to a different value from 
    `seq_along(df)`. It's useful to think of `i` as a pronoun, like "it".
    
    You might not have seen `seq_along()` before. It's a safe version of the 
    familiar `1:length(l)`.
    
1.  The __body__: `output[[i]] <- median(df[[i]])`. This is the code that does
    the work. It's run repeatedly, each time with a different value for `i`.
    The first iteration will run `output[[1]] <- median(df[[1]])`, 
    the second will run `output[[2]] <- median(df[[2]])`, and so on.

### Exercises

1.  Write for loops to:

    Think about the output, sequence, and body __before__ you start writing
    the loop.
    
    a. Compute the mean of every column in `mtcars`.
    
        ```{r}
        output <- vector("double", ncol(mtcars))
        for (i in seq_along(mtcars)) {
          output[[i]] <- mean(mtcars[[i]])
        }
        output
        
        mtcars %>% 
          summarise(across(.cols = dplyr::everything(),
                           mean,
                           .names = "mean_{.col}"))
        ```
    
    a. Determine the type of each column in `nycflights13::flights`.
    
        ```{r}
        flights <- nycflights13::flights
        output <- vector("list", ncol(flights))
        for (i in seq_along(flights)) {
          output[[i]] <- class(flights[[i]])
        }
        str(output)
        ```
    
    a. Compute the number of unique values in each column of `iris`.
    
        ```{r}
        output <- vector("integer", ncol(iris))
        for (i in seq_along(iris)) {
          output[[i]] <- n_distinct(iris[[i]])
        }
        output
        ```
    
    a. Generate 10 random normals from distributions with means of -10, 0, 10, and 100.
    
        ```{r}
        mean_vec <- c(-10, 0, 10, 100)
        output <- vector("list", length(mean_vec))
        for (i in seq_along(mean_vec)){
          output[[i]] <- rnorm(10, mean = mean_vec[i])
        }
        str(output)
        ```
    

1.  Eliminate the for loop in each of the following examples by taking 
    advantage of an existing function that works with vectors:
    
    ```{r, eval = FALSE}
    out <- ""
    for (x in letters) {
      out <- stringr::str_c(out, x)
    }
    out
    
    x <- sample(100)
    sd <- 0
    for (i in seq_along(x)) {
      sd <- sd + (x[i] - mean(x)) ^ 2
    }
    sd <- sqrt(sd / (length(x) - 1))
    
    x <- runif(100)
    out <- vector("numeric", length(x))
    out[1] <- x[1]
    for (i in 2:length(x)) {
      out[i] <- out[i - 1] + x[i]
    }
    ```
    
    ```{r}
    out <- ""
    for (x in letters) {
      out <- stringr::str_c(out, x)
    }
    out
    
    stringr::str_c(letters, collapse = "")
    ```
    
    ```{r}
        x <- sample(100)
        sd <- 0
        for (i in seq_along(x)) {
          sd <- sd + (x[i] - mean(x)) ^ 2
        }
        sd <- sqrt(sd / (length(x) - 1))
        sd
        
        sd(x)
    ```
    
    ```{r}
        x <- runif(100)
        out <- vector("numeric", length(x))
        out[1] <- x[1]
        for (i in 2:length(x)) {
          out[i] <- out[i - 1] + x[i]
        }
        out
        cumsum(x)
    ```
    
1.  Combine your function writing and for loop skills:

    a. Write a for loop that `prints()` the lyrics to the children's song 
       "Alice the camel".
       
       <pre>
       “Alice The Camel” Lyrics
        Alice the camel has five humps.
        Alice the camel has five humps.
        Alice the camel has five humps.
        So go, Alice, go!
        Boom, boom, boom, boom!
        
        Alice the camel has four humps.
        Alice the camel has four humps.
        Alice the camel has four humps.
        So go, Alice, go!
        Boom, boom, boom, boom!
        
        Alice the camel has three humps.
        Alice the camel has three humps.
        Alice the camel has three humps.
        So go, Alice, go!
        Boom, boom, boom, boom!
        
        Alice the camel has two humps.
        Alice the camel has two humps.
        Alice the camel has two humps.
        So go, Alice, go!
        Boom, boom, boom, boom!
        
        Alice the camel has one hump.
        Alice the camel has one hump.
        Alice the camel has one hump.
        So go, Alice, go!
        Boom, boom, boom, boom!
        
        Alice the camel has no humps.
        Alice the camel has no humps.
        Alice the camel has no humps.
        ‘Cause Alice is a horse, of course!
        </pre>

        ```{r}
          alice_song <- function(){
              times <- c("five", "four", "three", "two", "one", "no")
              song_lyrics <- vector("character", length(times))
          for (i in seq_along(times)) {
            if (times[i] == "no") {
              song_lyrics[[i]] <- str_glue(
              "\nAlice the camel has {times[i]} humps.
              Alice the camel has {times[i]} humps.
              Alice the camel has {times[i]} humps.
              'Cause Alice is a horse, of course!\n\n
              ")
              
            } else if (times[i] == "one") {
              song_lyrics[[i]] <- str_glue(
              "\nAlice the camel has {times[i]} hump.
              Alice the camel has {times[i]} hump.
              Alice the camel has {times[i]} hump.
              So go, Alice, go!
              Boom, boom, boom, boom!\n\n
              ")
            } else {
              song_lyrics[[i]] <- str_glue(
              "\nAlice the camel has {times[i]} humps.
              Alice the camel has {times[i]} humps.
              Alice the camel has {times[i]} humps.
              So go, Alice, go!
              Boom, boom, boom, boom!\n\n
              ")
            }
          }
          song_lyrics
        }
        
        song_lyrics <- alice_song()
        writeLines(str_c(song_lyrics, collapse = ""))
        ```


    a. Convert the nursery rhyme "ten in the bed" to a function. Generalise 
       it to any number of people in any sleeping structure.
       
        ```{r}
        roll_over <- function(num = "ten") {
          num_levels <-  c("one", "two", "three", "four", "five",
                       "six", "seven", "eight", "nine", "ten")
          num_fact <- factor(num, levels = num_levels)
          output <- vector("character", as.integer(num_fact))
          for (i in seq_along(output)){
            if(num_levels[[length(output)-(i-1)]] == "one"){
              output[[i]] <- str_glue(
              "There was {num_levels[[length(output)-(i-1)]]} in the bed",
               " and the little one said Ahhhhhh ...\n",
               "\n")
            } else {
              output[[i]] <- str_glue(
              "There were {num_levels[[length(output)-(i-1)]]} in the bed",
              " and the little one said roll over, roll over ...\n",
              "\n")
            }
          }
          output
        }
        writeLines(str_c(roll_over("three"), collapse = ""))
        writeLines(str_c(roll_over("five"), collapse = ""))
        ```
       
    a. Convert the song "99 bottles of beer on the wall" to a function.
       Generalise to any number of any vessel containing any liquid on 
       any surface.

1.  It's common to see for loops that don't preallocate the output and instead
    increase the length of a vector at each step:
    
    ```{r, eval = FALSE}
    output <- vector("integer", 0)
    for (i in seq_along(x)) {
      output <- c(output, lengths(x[[i]]))
    }
    output
    ```
    
    How does this affect performance? Design and execute an experiment.


<h2>For loop variations</h2>
## For loop variations

Here are some for loop variations.

1.  Modifying an existing object, instead of creating a new object.
1.  Looping over names or values, instead of indices.
1.  Handling outputs of unknown length.
1.  Handling sequences of unknown length.

### Modifying an existing object

Sometimes you want to use a for loop to modify an existing object. 

For example let's say we wanted to rescale an entire df.

To solve this with a for loop we again think about the three components:

1.  __Output__: we already have the output --- it's the same as the input!

1.  __Sequence__: we can think about a data frame as a list of columns, so 
    we can iterate over each column with `seq_along(df)`.

1.  __Body__: apply `rescale01()`.

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
df

rescale01 <- function(x){
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) /(rng[2] - rng[1])
}

for(i in seq_along(df)){
  df[[i]] <- rescale01(df[[i]])
}
df
```

### Looping patterns

To loop over a vector we have:

1. <span style="color: #008080;background-color:#9FDDBA">`for (i in seq_along(xs))`</span>,
and extracting the value with `x[[i]]`.

1. <span style="color: #008080;background-color:#9FDDBA">Loop over the elements:</span> 
`for (x in xs)`. Useful if you only care about side-effects, like plotting or 
printing.

1. <span style="color: #008080;background-color:#9FDDBA">Loop over the names</span>
`for (nm in names(xs))`. This gives you name, which you can use to access the 
value with `x[[nm]]`.

```{r, eval = FALSE}
results <- vector("list", length(x))
names(results) <- names(x)

for (i in seq_along(x)) {
  name <- names(x)[[i]]
  value <- x[[i]]
}
```


### Unknown output length

Sometimes you might not know how long the output will be.
<span style="color: #008080;background-color:#9FDDBA">Save results as
a list</span> and then compile into a single vector when loop is done.
used <span style="color: #008080;background-color:#9FDDBA">`unlist()`</span> to 
flatten a list of vectors into a single vector. A stricter option is to use
`purrr::flatten_dbl()`.

```{r}
means <- c(0,1,2)

output <- vector("list", length(means))
for (i in seq_along(means)){
  n <- sample(100, 1)
  output[[i]] <- rnorm(n, means[[i]])
}
str(output)
str(unlist(output))
```

This pattern occurs in other places too:

1.  You might be generating a long string. Instead of `paste()`ing together 
    each iteration with the previous, save the output in a character vector and
    then combine that vector into a single string with 
    `paste(output, collapse = "")`.
   
1.  You might be generating a big data frame. Instead of sequentially
    `rbind()`ing in each iteration, save the output in a list, then use 
    `dplyr::bind_rows(output)` to combine the output into a single
    data frame.

### Unknown sequence length

Sometimes you don't even know how long the sequence should run for -
you can use a while loop.

```{r, eval = FALSE}
while (condition) {
  # body
}
```

```{r}
flip <- function() sample(c("T", "H"), 1)
flips <- 0
nheads <- 0
while (nheads < 3) {
  if (flip() == "H") {
    nheads <- nheads + 1
  } else {
    nheads <- 0
  }
  flips <- flips + 1
}
flips
```

### Exercises

1.  Imagine you have a directory full of CSV files that you want to read in.
    You have their paths in a vector, 
    `files <- dir("data/", pattern = "\\.csv$", full.names = TRUE)`, and now
    want to read each one with `read_csv()`. Write the for loop that will 
    load them into a single data frame. 

1.  What happens if you use `for (nm in names(x))` and `x` has no names?
    What if only some of the elements are named? What if the names are
    not unique?

1.  Write a function that prints the mean of each numeric column in a data 
    frame, along with its name. For example, `show_mean(iris)` would print:
    
    ```{r, eval = FALSE}
    show_mean(iris)
    #> Sepal.Length: 5.84
    #> Sepal.Width:  3.06
    #> Petal.Length: 3.76
    #> Petal.Width:  1.20
    ```
    
    (Extra challenge: what function did I use to make sure that the numbers
    lined up nicely, even though the variable names had different lengths?)

1.  What does this code do? How does it work?

    ```{r, eval = FALSE}
    trans <- list( 
      disp = function(x) x * 0.0163871,
      am = function(x) {
        factor(x, labels = c("auto", "manual"))
      }
    )
    for (var in names(trans)) {
      mtcars[[var]] <- trans[[var]](mtcars[[var]])
    }
    ```

### Exercises

1.  Read the documentation for `apply()`. In the 2d case, what two for loops
    does it generalise?

1.  Adapt `col_summary()` so that it only applies to numeric columns
    You might want to start with an `is_numeric()` function that returns
    a logical vector that has a TRUE corresponding to each numeric column.
    
## The map functions

The {purrr} package provides a family of functions for the common task of
iteration. There is one function for each type of output:

* `map()`     makes a list.
* `map_lgl()` makes a logical vector.
* `map_int()` makes an integer vector.
* `map_dbl()` makes a double vector.
* `map_chr()` makes a character vector.

Each function takes a vector as input, applies a function to each piece, and then returns a new vector that's the same length (and has the same names) as the input. The type of the vector is determined by the suffix to the map function. 

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

output <- vector("double", length(df))
for (i in seq_along(df)) {
  output[[i]] <- mean(df[[i]])
}
output

# generlise it into a func
col_mean <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- mean(df[[i]])
  }
  output
}

# oh-oh two more functions of a slight variation needed!
col_median <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- median(df[[i]])
  }
  output
}
col_sd <- function(df) {
  output <- vector("double", length(df))
  for (i in seq_along(df)) {
    output[i] <- sd(df[[i]])
  }
  output
}
col_mean(df)
col_sd(df)
col_median(df)
```

We would then generalise the function after realising we can pass functions
into functions as an argument!!`r emo::ji("flex")`

```{r}
col_summary <- function(df, fun) {
  out <- vector("double", length(df))
  for (i in seq_along(df)) {
    out[i] <- fun(df[[i]])
  }
  out
}
col_summary(df, median)
col_summary(df, mean)
```

With {purrr} functions we could do this in a synch! `r emo::ji("cat")`

```{r}
map_dbl(df, mean)
map_dbl(df, median)
# or using pipes
df %>% map_dbl(mean)
df %>% map_dbl(median)
```


There are a few differences between `map_*()` and the `col_summary()` function
we wrote:

*   {purrr} functions are implemented in C - faster.
    
*   The second argument, `.f`, the function to apply, can be:

    - a formula, 
    - a character vector, or 
    - an integer vector. 
    
*   `map_*()` uses ... ([dot dot dot]) to pass along additional arguments 
    to `.f` each time it's called:

    ```{r}
    map_dbl(df, mean, trim = 0.5)
    ```

*   The map functions also preserve names:

    ```{r}
    z <- list(x = 1:3, y = 4:5)
    map_int(z, length)
    ```

### Shortcuts

There are a few shortcuts that you can use with `.f`. Say we want to fit a 
linear model to each group in a dataset.

```{r}
(models <- mtcars %>% 
  split(.$cyl) %>% 
   # explicitly create a func, with an input which is each
   # split
  map(function(df) lm(mpg ~ wt, data = df)))
```

purrr provides a convenient shortcut for an anonymous function:
a one-sided formula.

```{r}
(models <- mtcars %>% 
  split(.$cyl) %>% 
  map(~lm(mpg ~ wt, data = .)))
```

We us `.` to refer to the current list element.

Let's say we want to extract a summary statistic like the $R^2$. 

- First run `summary()` and 
- Then extract the component called `r.squared`. 

```{r}
models %>% 
  map(summary) %>% 
  map_dbl(~.$r.squared) # alternate would be: map_dbl(function(df) df$r.squared)
```

__But__ extracting named components is a common operation, so purrr provides 
an even shorter shortcut: you can use a string. `r emo::ji("grin")`

```{r}
models %>% 
  map(summary) %>% 
  map_dbl("r.squared")
```

You can also use an integer to select elements by position: 

```{r}
x <- list(list(1, 2, 3), list(4, 5, 6), list(7, 8, 9))
x %>% map_dbl(2)
```

### Exercises

1.  Write code that uses one of the map functions to:

    a. Compute the mean of every column in `mtcars`.
    
        ```{r}
        mtcars %>% 
          map_dbl(mean)
        ```
    
    a. Determine the type of each column in `nycflights13::flights`.
    
        ```{r}
        nycflights13::flights %>% 
          map(class) %>% 
          str()
        ```
    
    a. Compute the number of unique values in each column of `iris`.
    
      ```{r}
      iris %>% 
        map_int(n_distinct)
      ```
    
    a. Generate 10 random normals from distributions with means of -10, 0, 10, and 100.
    
      ```{r}
      means <- c(-10,0,10,100) 
      means %>% 
        map(.f = rnorm, n = 10) %>% 
        str()
      ```
    
1.  How can you create a single vector that for each column in a data frame
    indicates whether or not it's a factor?
    
    ```{r}
    library(palmerpenguins)
    penguins %>% 
      map_chr(is.factor)
    ```
    
1.  What happens when you use the map functions on vectors that aren't lists?
    What does `map(1:5, runif)` do? Why?
    
    It passes 1, then 2, then 3 etc, to the first argument of `runif` which
    is `n` - how many numbers you want the random uniform function to generate.
    
    ```{r}
    map(1:5, runif)
    ```
    
    
1.  What does `map(-2:2, rnorm, n = 5)` do? Why?
    What does `map_dbl(-2:2, rnorm, n = 5)` do? Why?
    
     What does `map(-2:2, rnorm, n = 5)` do? Why?
     It send -2, -1, 0, 1, to rnorm as the mean value and returns
     a list.
     
     What does `map_dbl(-2:2, rnorm, n = 5)` do? Why? 
     It send -2, -1, 0, 1, 2 to rnorm as the mean value and returns
     a double vector the same size as the input. But uh oh we gave it the
     mean vector of size 5 but this is used to generate 5 vectors of
     n = 5 in length. This is unexpected and hence will error.
     
      ```{r}
      map(-2:2, rnorm, n = 5)
      ```
      
    

1.  Rewrite `map(x, function(df) lm(mpg ~ wt, data = df))` to eliminate the 
    anonymous function. 
    
    ```{r, eval = FALSE}
    # map(x, function(df) lm(mpg ~ wt, data = df))
    
    map(x, ~lm(mpg~wt, data = .))
    ```
    
