---
title: "Chapter 1 - Data Viz with ggplot"
author: "Data Analytics"
date: "10/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

### Do cars with big engines use more fuel than cars with small engines?

Hypothesis: Cars with bigger engines use more fuel, i.e. the fuel efficiency declines as the engine size gets bigger. If `miles per gallon` was on the y-axis and `engine size` on the x-axis we would see a decreasing trend.

```{r}
ggplot2::mpg

# create coordinate system
ggplot(data = mpg, aes(x = displ,
                       y = hwy))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ,
                           y = hwy))
```

My hypothesis has been confirmed.

```{r}
num_rows <- nrow(mtcars)
num_cols <- ncol(mtcars)
ex4_plot <- ggplot(data = mpg,
                   aes(x = hwy,
                       y = cyl)) +
  geom_point()
```

#### Exercises
1. Run `ggplot(data = mpg)`. What do you see? <br>
Ans: An empty canvas of a plot. If you add the `aes(x = xx, y = yy)` you will see an empty canvas with the axes drawn.
1. How many `rows` are there in `mtcars`? Columns? <br>
Ans: Number of rows is `r num_rows`, cols is `r num_cols`.
1. What does the `drv` variable describe?
Ans: 'The type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd'
1. Make a scatterplot of hwy versus cyl.

```{r}
ex4_plot
```

1. What happens if you make a scatterplot of `class` versus `drv`? Why is this plot not useful?

```{r}
ggplot(data = mpg, aes(x = class, 
                       y = drv)) +
  geom_point()
```

These are 2 categorical variables here so this isn't very useful.



