---
title: "Chapter 1 - Data Viz with ggplot"
author: "Data Analytics"
date: "10/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(tidyverse)
library(flair)
```

### Do cars with big engines use more fuel than cars with small engines?

Hypothesis: Cars with bigger engines use more fuel, i.e. the fuel efficiency declines as the engine size gets bigger. If `miles per gallon` was on the y-axis and `engine size` on the x-axis we would see a decreasing trend.

```{r}
ggplot2::mpg

# create coordinate system
ggplot(data = mpg, aes(x = displ,
                       y = hwy))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ,
                           y = hwy))
```

My hypothesis has been confirmed.

```{r}
num_rows <- nrow(mtcars)
num_cols <- ncol(mtcars)
ex4_plot <- ggplot(data = mpg,
                   aes(x = hwy,
                       y = cyl)) +
  geom_point()
```

### Aesthetics

#### Exercises
1. Run `ggplot(data = mpg)`. What do you see? <br>
Ans: An empty canvas of a plot. If you add the `aes(x = xx, y = yy)` you will see an empty canvas with the axes drawn.
1. How many `rows` are there in `mtcars`? Columns? <br>
Ans: Number of rows is `r num_rows`, cols is `r num_cols`.
1. What does the `drv` variable describe?
Ans: 'The type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd'
1. Make a scatterplot of hwy versus cyl.

```{r}
ex4_plot
```

1. What happens if you make a scatterplot of `class` versus `drv`? Why is this plot not useful?

```{r}
ggplot(data = mpg, aes(x = class, 
                       y = drv)) +
  geom_point()
```

These are 2 categorical variables here so this isn't very useful.


You can give extra information about your dataset by mapping data to aesthetics like `size`, `colour`, `shape`.

```{r aes1, include=FALSE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, 
                           colour = class))
```

```{r, echo = FALSE}
decorate("aes1") %>% 
  flair("colour = class", background = "#9FDDBA", 
        color = "#008080")
```

When you put a feature against an aesthetic `ggplot` will assign a unique `level` (here <span style="color: #008080;background-color:#9FDDBA">colour</span> to each <span style="color: #008080;background-color:#9FDDBA">class</span> of the feature) -> this process is called `scaling`.

```{r size_aes, include=FALSE}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy,
                 size = class))
```

```{r, echo=FALSE}
decorate("size_aes") %>% 
  flair("size = class", background = "#800000", 
        color = "#FFE4E1")
```

Mapping a unordered variable like `class` to an <span style="color: #FFE4E1;background-color:#800000">ordered aesthetic</span> like `size` is not a good idea, and we get a <span style="color: #FFE4E1;background-color:#800000">warning</span> here.

```{r alpha_aes, include=FALSE}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy,
                 alpha = class))
```

```{r, echo=FALSE}
decorate("alpha_aes") %>% 
  flair("alpha = class", background = "#9FDDBA", 
        color = "#008080")
```

<span style="color: #008080;background-color:#9FDDBA">`alpha`</span> is another aesthetic that we can use.

```{r shape_aes, include=FALSE, warning=TRUE}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy,
                 shape = class))
```

```{r, echo=FALSE}
decorate("shape_aes") %>% 
  flair("shape = class", background = "#800000", 
        color = "#FFE4E1")
```

`ggplot` only uses **6** <span style="color: #FFE4E1;background-color:#800000">`shape`</span>s at a time, so we're missing `suv`! When mapping to aesthetics think carefully about which aesthetic makes sense.

If you look at the `colour = class` plot vs the `shape = class` plot you may well think - it looks like previously labelled `suv` is now being considered to be `pickup`!? There are many points and the points get plotted on top of each other - if you look near `displ = 5 and hwy = 12` notice that there is an `pickup` point there ... previously there was an `suv` point there. Let's jitter the data so that the points lie a little away from each other to show that there's not any issue with `ggplot`, instead the data lying on top of each other shows the last `class`.

```{r}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy,
                 colour = class))

ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy,
                 shape = class))

ggplot(data = mpg) +
  geom_jitter(aes(x = displ, y = hwy,
                 colour = class))

ggplot(data = mpg) +
  geom_jitter(aes(x = displ, y = hwy,
                 shape = class))
```

Occassionally you may want to change all the points sizes uniformly (irrespective of any other data feature) - in this case you may put the aesthetic on the outside of the `aes()`.

```{r uniform_col, include=FALSE}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy),
             colour = "blue")
```

```{r, echo = FALSE}
decorate('uniform_col') %>% 
  flair('colour = "blue"', 
        background = "#e5989b", 
        color = "#6d6875")

```

```{r uniform_size, include=FALSE}
ggplot(data = mpg) +
  geom_point(aes(x = displ, 
                 y = hwy),
             size = 4)
```

```{r, echo = FALSE}
decorate('uniform_size') %>% 
  flair("size = 4", 
        background = "#e5989b", 
        color = "#6d6875")

```

#### Exercises

1. What’s gone wrong with this code? Why are the points not blue?

```{r ex1, include=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, 
                           y = hwy, color = "blue"))
```

```{r, echo=FALSE}
decorate("ex1") %>% 
  flair('color = "blue"',
        background = "#800000", 
        color = "#FFE4E1")
```


The <span style="color: #FFE4E1;background-color:#800000">`color`</span> attribute maps to a static colour and yet appears within `aes()`.

To correct, put `colour = 'blue'` outside the `aes()`

```{r ex1_correct, include=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, 
                           y = hwy), 
             color = "blue")
```

```{r, echo=FALSE}
decorate("ex1_correct") %>% 
  flair('color = "blue"',
        background = "#e5989b", 
        color = "#6d6875")
```

2. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

`manufacturer`, `model`, `trans`, `drv`, `fl` and `class` are categories (I also think `cyl` may be considered a category since it takes on 4 values based on a limited number of cylinders); `displ`, `year`, `cty`, `hwy`, `cyl` are continuous. You can convert `mpg` to a tibble (if it is not already) and check the types using `glimpse()`.

```{r}
glimpse(mpg)
```

3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy, 
                 colour = cyl))

ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy, 
                 colour = cty))
```

The colours are scaled from the lowest to the highest value using dark to light blue.

4. What happens if you map the same variable to multiple aesthetics?

```{r ex4_gg, include = FALSE}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy,
                 colour = class,
                 size = class))
```

```{r, echo=FALSE}
decorate("ex4_gg") %>% 
  flair("colour = class",
        background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("size = class",
        background = "#9FDDBA", 
        color = "#008080")
```

Here we have both <span style="color: #008080;background-color:#9FDDBA">`colour` and `size`</span> mapped to `class`. The plot offers no new information with the second aesthetic so I'd consider it a poor plot. You should use aesthetics mapped to different features in your data to meet with the **Axiom** that _'The greatest value of a picture is when it forces us to notice what we never expected to see. --John Tukey'_ as mentioned in the beginning of the __Aesthetic Mappings__ section!

5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)

__Tip: Did you know that when you call `?geom_point` you can go to an example and highlight it, press 'Ctrl + Enter' and it will place that example in your console and run it?__

<img src = assets/help_ctrl_enter.PNG width = 250px>

<img src = assets/ctrl_enter_example_help_file2.gif>

```{r}
# For shapes that have a border (like 21), you can colour the inside and
# outside separately. Use the stroke aesthetic to modify the width of the
# border
ggplot(mtcars, aes(wt, mpg)) +
  geom_point(shape = 2, colour = "black", fill = "white", size = 5, stroke = 5)

ggplot(mtcars, aes(wt, mpg)) +
  geom_point(shape = 23, colour = "black", fill = "white", size = 5, stroke = 3)
```

<img src = assets/shape.PNG>

Stroke changes the border width as described in the example comment, and works with any shape that has a border hence will work with shapes `0-14`, and `21-24`. An example of each is above.

6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)? Note, you’ll also need to specify x and y.

```{r ex6_gg, include = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, 
            colour = displ < 5))
```

```{r, echo=FALSE}
decorate("ex6_gg") %>% 
  flair("colour = displ < 5",
        background = "#9FDDBA", 
        color = "#008080")
```

Points are split according to the criteria <span style="color: #008080;background-color:#9FDDBA">`displ < 5`</span> and are coloured differently depending on what value they have for displacement.

### Facets
Another way to add information on your plot is to use `facets`. 

- `facet_wrap()` allows faceting by a single variable

  * First argument is a formula `~ var_name`
  * Variable passed in should be discrete.
  
- `facet_grid()` allows the faceting by 2 variables.

  * `var_name1 ~ var_name2`
  
```{r facet1, include=FALSE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ,
                           y = hwy)) +
  facet_wrap(~ class, nrow = 2)
```

```{r, echo=FALSE}
decorate("facet1") %>% 
  flair("facet_wrap(~ class, nrow = 2)", 
        background = "#9FDDBA", 
        color = "#008080")
```

```{r facet2, include=FALSE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ,
                           y = hwy)) +
  facet_grid(drv ~ class)
```

```{r, echo=FALSE}
decorate("facet2") %>% 
  flair("facet_grid(drv ~ class)", 
        background = "#9FDDBA", 
        color = "#008080")
```

#### Exercises

1.  What happens if you facet on a continuous variable?

```{r facet_ex1, include=FALSE}
ggplot(data = mpg) +
  geom_point(aes(x = displ,
                 y = hwy)) +
  facet_wrap(~ cty)
```

```{r, echo=FALSE}
decorate("facet_ex1") %>% 
  flair("facet_wrap(~ cty)", 
        background = "#800000", 
        color = "#FFE4E1")
```

    The plot is created but it doesn't make sense to split our data 
    by a continuous variable. 

1. What do the empty cells in plot with `facet_grid(drv ~ cyl)` mean? How do they relate to this plot?

    The empty spots refer to that `drv and cyl combination` being missing.
    
    ```
    ggplot(data = mpg) + <br>
      geom_point(mapping = aes(x = drv, y = cyl))
      
    ```  
    
    ```{r facet_ex2}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = drv, y = cyl))
    ```

    It relates to the above in that you can see that the 
    `drv = 4; cyl = 5` combination has no observations. 
    This is also the case in the faceted plot.

1. What plots does the following code make? What does . do?

    ```{r fct3_1, include=FALSE}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy)) +
      facet_grid(drv ~ .)
    
    ```
    
    ```{r fct3_2, include=FALSE}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy)) +
      facet_grid(. ~ cyl)
    ```
    
    ```{r, echo=FALSE}
    decorate("fct3_1") %>% 
      flair("facet_grid(drv ~ .)", 
            background = "#9FDDBA", 
            color = "#008080")
    ```

    The <span style="color: #008080;background-color:#9FDDBA">`facet_grid
    (drv ~ .)`</span> facets the `drv` categories into rows. It says 
    facet this plot by `drv` as row panels but I don't want anything 
    as column panels (hence the `.`).
  
    ```{r, echo=FALSE}
    decorate("fct3_2") %>% 
      flair("facet_grid(. ~ cyl)", 
            background = "#9FDDBA", 
            color = "#008080")
    ```

    The <span style="color: #008080;background-color:#9FDDBA">`facet_grid(. ~ cyl)`</span> facets the `cyl` categories into columns. It says facet this plot by `cyl` as column panels but I don't want anything as row panels (hence the `.`).
    
1. Take the first faceted plot in this section:

    ```
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy)) + 
      facet_wrap(~ class, nrow = 2)
    
    ```
    What are the advantages to using faceting instead of the colour aesthetic? What are the disadvantages? How might the balance change if you had a larger dataset?
    
    ```{r}
    ggplot(data = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy)) + 
      facet_wrap(~ class, nrow = 2)
    ```
    
    * What are the advantages to using faceting instead of the colour aesthetic? <br>
    You can see the patterns in each class much easier.

    * What are the disadvantages? <br>
    You may lose sight of the overall trend across all observations, and comparing categories becomes more "work".

    * How might the balance change if you had a larger dataset? <br>
    If you have many categories it becomes overwhelming.

5. Read ?facet_wrap. What does nrow do? What does ncol do? What other options control the layout of the individual panels? Why doesn’t facet_grid() have nrow and ncol arguments? <br>
From the help page these are the main aspects that alter your visuals.

* `nrow` and `ncol` specify the number of rows or columns you're looking for in your plot.
* `scales`: Should scales be fixed ("fixed", the default), free ("free"), or free in one dimension ("free_x", "free_y")?
* Use the `labeller` option to control how labels are printed.
* Use `strip.position` to display the facet labels at the side of your choice. Setting it to `bottom` makes it act as a subtitle for the axis. This is typically used with free scales and a theme without boxes around strip labels.

6. When using facet_grid() you should usually put the variable with more unique levels in the columns. Why? <br>
I would think this orientation makes best use of the screen real estate available.

### Geometric Objects

```{r, echo = FALSE, out.width = "50%", fig.align="default", message = FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

```{r plots, echo = FALSE, out.width = "50%", fig.align="default", message = FALSE, include=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

```{r, echo=FALSE}
decorate("plots") %>% 
  flair("geom_point",
        background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("geom_smooth",
        background = "#9FDDBA", 
        color = "#008080")

```


The above plots are made using the same data, but the <span style="color: #008080;background-color:#9FDDBA">`geoms`</span> are different. To change the plot type i.e. `bar`, `boxplot`, `line` etc. you just <span style="color: #008080;background-color:#9FDDBA">change the `geom` you add to `ggplot`.</span> The `aesthetics` that work with each `geom` is different however. E.g. you may set the <span style="color: #008080;background-color:#9FDDBA">`linetype`</span> of a line, but you __can't__ change the <span style="color: #FFE4E1;background-color:#800000">`shape` of a line.</span>

```{r linetype1, include=FALSE}
ggplot(data = mpg) +
  geom_line(aes(x = displ, y = hwy, linetype = drv))
```

```{r, echo=FALSE}
decorate("linetype1") %>% 
  flair("linetype = drv",
        background = "#9FDDBA", 
        color = "#008080")
```


```{r linetype2, include=FALSE}
ggplot(data = mpg) +
  geom_smooth(aes(x = displ, y = hwy, linetype = drv))
```

```{r, echo=FALSE}
decorate("linetype2") %>% 
  flair("linetype = drv",
        background = "#9FDDBA", 
        color = "#008080")
```

```{r two_geoms1, include=FALSE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, colour = drv)) +
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, colour = drv))

```

```{r, echo=FALSE}
decorate("two_geoms1") %>% 
  flair("geom_point(mapping = aes(x = displ, y = hwy, colour = drv))",
        background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv, colour = drv))",
        background = "#9FDDBA", 
        color = "#008080")
```

We are also able to add more than <span style="color: #008080;background-color:#9FDDBA">one geom</span> on a plot. In the above we show both the <span style="color: #008080;background-color:#9FDDBA">raw data points</span> coloured by the drive type - e.g. 4 wheel, rear drive or front wheel, as well as each drive types <span style="color: #008080;background-color:#9FDDBA">smoothing line</span>. But the downside to the above is that I had to specify the `mapping` twice with many repeated elements. To avoid this `ggplot` provides a handy alternate - you may set the `mapping` in the `ggplot()` call itself - the `mapping` is treated as __global__ to all subsequent `geoms` unless you specifically override these.

```{r two_geoms2, include=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, colour = drv)) +
  geom_point() +
  geom_smooth(mapping = aes(linetype = drv))
```

```{r, echo = FALSE}
decorate("two_geoms2") %>% 
  flair("mapping = aes(x = displ, y = hwy, colour = drv)", 
        background = "#9FDDBA", 
        color = "#008080") 
```

Also in the above you may have noticed that <span style="color: #008080;background-color:#9FDDBA">`geom_smooth()`</span> included it's own mapping. This basically says _'Hey, take all the attributes as per the ggplot mapping BUT replace these components (if it exists) as described in this geom'._

```{r spec_override1, include=FALSE}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(colour = class)) +
  geom_smooth()
```

```{r, echo = FALSE}
decorate("spec_override1") %>% 
  flair("mapping = aes(x = displ, y = hwy)",
        background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("mapping = aes(colour = class)",
        background = "#9FDDBA", 
        color = "#008080")
```
For example here when it gets to the <span style="color: #008080;background-color:#9FDDBA">__geom_point(mapping = aes(colour = class))__</span> line it asks:

1. Does this mapping specify a new `x`? It doesn't, so `x = displ` is set, as per the `ggplot()` `mapping()`.
1. Does this mapping specify a new `y`? It doesn't, so `y = hwy` is set, as per the `ggplot()` `mapping()`.
1. Is there anything extra set? Yip, `colour = class` is set, so let me apply that to this `geom_point`'s only, hence the scatterpoint plots are coloured by the `class` of the vehicle.

You may also __even__ change the __data__ for each layer! Here the `smoothing` function is included but for a subset of the data.

```{r spec_override2, include=FALSE}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(mapping = aes(colour = class)) +
  geom_smooth(
    data = filter(mpg, class == "subcompact"),
    se = FALSE # remove the error bands that display around the line
  )
```

```{r, echo = FALSE}
decorate("spec_override2") %>% 
  flair('data = filter(mpg, class == "subcompact")', 
        background = "#9FDDBA", 
        color = "#008080")
```

#### Exercises

1.  What geom would you use to draw a line chart? A boxplot? 
    A histogram? An area chart?
    
    * line chart: geom_line()
    * boxplot: geom_boxplot()
    * histogram: geom_histogram()
    * Area chart: geom_area()
    
1.  Run this code in your head and predict what the output will look like.
    Then, run the code in R and check your predictions.
    
    ```{r, eval = FALSE}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
      geom_point() + 
      geom_smooth(se = FALSE)
    ```
    In my head this will produce a scatterplot with `displ` on the x-axis,
    `hwy` on the y-axis where each point is coloured by the 
    type of `drive` the vehicle is (4 wheel, front-wheel, rear-wheel). 
    The plot will also contain smoothing lines (with no error bands) 
    for each drive type (since `colour` is set in the `ggplot` 
    mapping layer, and is applicable for both points, and the 
    smoothing lines).
    
    ```{r, eval = TRUE}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
      geom_point() + 
      geom_smooth(se = FALSE)
    ```    
1.  What does `show.legend = FALSE` do?  What happens if you remove it?  
    Why do you think I used it earlier in the chapter?   
    
    It removes the legend that shows up when we add certain aesthetics
    such as `colour`, `shape` etc.
    
    If we remove it the legend will show by the fact that a certain non-xy 
    aesthetic has been added.
    
    It was used to remove the legend earlier which would have shown up due to
    the `colour = drv` aesthetic.

1.  What does the `se` argument to `geom_smooth()` do?
    It sets the errors bands on the smoothing function to either on or off.
    

1.  Will these two graphs look different? Why/why not?

    ```{r, eval = FALSE}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point() + 
      geom_smooth()
    
    ggplot() + 
      geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
    ```
    They will look the same. In the first plot the `data = mpg, mapping = aes(x = displ, y = hwy)`
    will be inherited by both `geom_point()` and `geom_smooth()`.
    In the second plot the `data = mpg, mapping = aes(x = displ, y = hwy)` is repeated 
    in each geom and hence both plots will be the same.
    
    ```{r, out.width = "50%", fig.align="default", message = FALSE}
    ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_point() + 
      geom_smooth()
    
    ggplot() + 
      geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
      geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
    ```
    
1.  Recreate the R code necessary to generate the following graphs.
        
    ```{r}
    ggplot(data = mpg, aes(x = displ, y = hwy)) +
      geom_point(size = 2) +
      geom_smooth(se = FALSE)
    ```
    ```{r}
    ggplot(data = mpg, aes(x = displ, y = hwy)) +
      geom_point(size = 2) +
      geom_smooth(aes(group = drv), se = FALSE)
    ```
    ```{r}
    ggplot(data = mpg, aes(x = displ, y = hwy, colour = drv)) +
      geom_point(size = 2) +
      geom_smooth(se = FALSE)
    ```
    ```{r}
    ggplot(data = mpg, aes(x = displ, y = hwy)) +
      geom_point(aes(colour = drv)) +
      geom_smooth(se = FALSE)
    ```
    ```{r}
    ggplot(data = mpg, aes(x = displ, y = hwy)) +
      geom_point(aes(colour = drv)) +
      geom_smooth(aes(linetype = drv), se = FALSE)
    ```
    ```{r}
    ggplot(data = mpg, aes(x = displ, y = hwy)) +
      geom_point(size = 3, colour = "white") +
      geom_point(aes(colour = drv))
    ```





































