---
title: "Chapter 1 - Data Viz with ggplot"
author: "Data Analytics"
date: "10/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(tidyverse)
library(flair)
```

#### Do cars with big engines use more fuel than cars with small engines?

Hypothesis: Cars with bigger engines use more fuel, i.e. the fuel efficiency declines as the engine size gets bigger. If `miles per gallon` was on the y-axis and `engine size` on the x-axis we would see a decreasing trend.

```{r}
ggplot2::mpg

# create coordinate system
ggplot(data = mpg, aes(x = displ,
                       y = hwy))

ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ,
                           y = hwy))
```

My hypothesis has been confirmed.

```{r}
num_rows <- nrow(mtcars)
num_cols <- ncol(mtcars)
ex4_plot <- ggplot(data = mpg,
                   aes(x = hwy,
                       y = cyl)) +
  geom_point()
```

##### Exercises
1. Run `ggplot(data = mpg)`. What do you see? <br>
Ans: An empty canvas of a plot. If you add the `aes(x = xx, y = yy)` you will see an empty canvas with the axes drawn.
1. How many `rows` are there in `mtcars`? Columns? <br>
Ans: Number of rows is `r num_rows`, cols is `r num_cols`.
1. What does the `drv` variable describe?
Ans: 'The type of drive train, where f = front-wheel drive, r = rear wheel drive, 4 = 4wd'
1. Make a scatterplot of hwy versus cyl.

```{r}
ex4_plot
```

1. What happens if you make a scatterplot of `class` versus `drv`? Why is this plot not useful?

```{r}
ggplot(data = mpg, aes(x = class, 
                       y = drv)) +
  geom_point()
```

These are 2 categorical variables here so this isn't very useful.


You can give extra information about your dataset by mapping data to aesthetics like `size`, `colour`, `shape`.

```{r aes1, include=FALSE}
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy, 
                           colour = class))
```

```{r, echo = FALSE}
decorate("aes1") %>% 
  flair("colour = class", background = "#9FDDBA", 
        color = "#008080")
```

When you put a feature against an aesthetic `ggplot` will assign a unique `level` (here <span style="color: #008080;background-color:#9FDDBA">colour</span> to each <span style="color: #008080;background-color:#9FDDBA">class</span> of the feature) -> this process is called `scaling`.

```{r size_aes, include=FALSE}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy,
                 size = class))
```

```{r, echo=FALSE}
decorate("size_aes") %>% 
  flair("size = class", background = "#800000", 
        color = "#FFE4E1")
```

Mapping a unordered variable like `class` to an <span style="color: #FFE4E1;background-color:#800000">ordered aesthetic</span> like `size` is not a good idea, and we get a <span style="color: #FFE4E1;background-color:#800000">warning</span> here.

```{r alpha_aes, include=FALSE}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy,
                 alpha = class))
```

```{r, echo=FALSE}
decorate("alpha_aes") %>% 
  flair("alpha = class", background = "#9FDDBA", 
        color = "#008080")
```

<span style="color: #008080;background-color:#9FDDBA">`alpha`</span> is another aesthetic that we can use.

```{r shape_aes, include=FALSE, warning=TRUE}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy,
                 shape = class))
```

```{r, echo=FALSE}
decorate("shape_aes") %>% 
  flair("shape = class", background = "#800000", 
        color = "#FFE4E1")
```

`ggplot` only uses **6** <span style="color: #FFE4E1;background-color:#800000">`shape`</span>s at a time, so we're missing `suv`! When mapping to aesthetics think carefully about which aesthetic makes sense.

If you look at the `colour = class` plot vs the `shape = class` plot you may well think - it looks like previously labelled `suv` is now being considered to be `pickup`!? There are many points and the points get plotted on top of each other - if you look near `displ = 5 and hwy = 12` notice that there is an `pickup` point there ... previously there was an `suv` point there. Let's jitter the data so that the points lie a little away from each other to show that there's not any issue with `ggplot`, instead the data lying on top of each other shows the last `class`.

```{r}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy,
                 colour = class))

ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy,
                 shape = class))

ggplot(data = mpg) +
  geom_jitter(aes(x = displ, y = hwy,
                 colour = class))

ggplot(data = mpg) +
  geom_jitter(aes(x = displ, y = hwy,
                 shape = class))
```

Occassionally you may want to change all the points sizes uniformly (irrespective of any other data feature) - in this case you may put the aesthetic on the outside of the `aes()`.

```{r uniform_col, include=FALSE}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy),
             colour = "blue")
```

```{r, echo = FALSE}
decorate('uniform_col') %>% 
  flair('colour = "blue"', 
        background = "#e5989b", 
        color = "#6d6875")

```

```{r uniform_size, include=FALSE}
ggplot(data = mpg) +
  geom_point(aes(x = displ, 
                 y = hwy),
             size = 4)
```

```{r, echo = FALSE}
decorate('uniform_size') %>% 
  flair("size = 4", 
        background = "#e5989b", 
        color = "#6d6875")

```

##### Exercises
1. What’s gone wrong with this code? Why are the points not blue?

```{r ex1, include=FALSE}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, 
                           y = hwy, color = "blue"))
```

```{r, echo=FALSE}
decorate("ex1") %>% 
  flair('color = "blue"',
        background = "#800000", 
        color = "#FFE4E1")
```


The <span style="color: #FFE4E1;background-color:#800000">`color`</span> attribute maps to a static colour and yet appears within `aes()`.

1. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

`manufacturer`, `model`, `cyl`, `trans`, `drv`, `fl` and `class` are categories; `displ`, `year`, `cty`, `hwy` are continuous. You can convert `mpg` to a tibble (if it is not already) and check the types using `glimpse()`.

```{r}
glimpse(mpg)
```

1. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

```{r}
ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy, 
                 colour = cyl))

ggplot(data = mpg) +
  geom_point(aes(x = displ, y = hwy, 
                 colour = cty))
```

The colours are kinda scaled from the lowest to the highest.






