---
title: "Chapter 21 - R Markdown"
author: "Vebash Naidoo"
date: "25/10/2020"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE,
                      comment = "#>", collapse = TRUE)
```

```{r libs}
options(scipen=10000)
library(tidyverse)
library(flair)
library(emo)
library(lubridate)
library(magrittr)
library(tidyquant)
theme_set(theme_tq())
```

### Overview of commands

```{r cat1, echo = FALSE, comment = ""}
cat(readr::read_file("docs/assets/markdown.Rmd"))
```

<br>

```{r cat_cv, echo = FALSE, comment = ""}
cat(readr::read_file("docs/mock_cv.Rmd"))
```

[Mock CV](mock_cv.html) <br>
<br>

### Options

The following table summarises which types of output each <span style="color: #008080;background-color:#9FDDBA">option suppresses</span>:

Option             | Run code | Show code | Output | Plots | Messages | Warnings 
-------------------|----------|-----------|--------|-------|----------|---------
`eval = FALSE`     | -        |           | -      | -     | -        | -
`include = FALSE`  |          | -         | -      | -     | -        | -
`echo = FALSE`     |          | -         |        |       |          |
`results = "hide"` |          |           | -      |       |          | 
`fig.show = "hide"`|          |           |        | -     |          |
`message = FALSE`  |          |           |        |       | -        |
`warning = FALSE`  |          |           |        |       |          | -


<br> 

The code below generates Table \@ref(tab:kable).

```{r}
knitr::kable(
  mtcars[1:5, ],
  caption = "A knir table"
)
```

### Caching

Each knit of a document starts from a <span style="color: #008080;background-color:#9FDDBA">completely clean slate</span>. 

- Pro: great for <span style="color: #008080;background-color:#9FDDBA">reproducibility</span>.
- Con: It can be <span style="color: #FFE4E1;background-color:#E3242B">painful</span> if you have some computations that take a long time. 
- The solution is <span style="color: #008080;background-color:#9FDDBA">`cache = TRUE`</span>. When set, the output of the chunk is saved to a specially named file on disk. On subsequent runs, knitr will check to see if the code has changed, and if it hasn't, it will reuse the cached results.
- Note: The caching system must be used with care, because by default it is based on the code only, not its dependencies. For example, here the `processed_data` chunk depends on the `raw_data` chunk:
  
  ````markdown
  
      `r ''````{r raw_data}
      rawdata <- readr::read_csv("a_very_large_file.csv")
      `r ''````
      
      `r ''````{r processed_data, cache = TRUE}
      processed_data <- rawdata %>% 
        filter(!is.na(import_var)) %>% 
        mutate(new_variable = complicated_transformation(x, y, z))
      `r ''````
  ````

    Caching the `processed_data` chunk means that it will get re-run 
    if the dplyr pipeline is changed, but it won't get rerun if 
    the `read_csv()` call changes. You can avoid that problem with the 
    <span style="color: #008080;background-color:#9FDDBA">`dependson`</span>
    chunk option:

  ````markdown
      `r ''````{r processed_data, cache = TRUE, dependson = "raw_data"}
      processed_data <- rawdata %>% 
        filter(!is.na(import_var)) %>% 
        mutate(new_variable = complicated_transformation(x, y, z))
      `r ''````
  ````

    `dependson` should contain a character vector of *every* chunk 
    that the cached chunk depends on.    


- Note that the chunks won't update if `a_very_large_file.csv` changes, because knitr caching only tracks changes within the `.Rmd` file. Use the <span style="color: #008080;background-color:#9FDDBA">`cache.extra`</span>  option to track changes to the file. The R expression will invalidate the cache whenever it changes. A good function to use is <span style="color: #008080;background-color:#9FDDBA">`file.info()`</span>: it returns a bunch of information about the file including when it was last modified. 

  ````markdown
      `r ''````{r raw_data, cache.extra = file.info("a_very_large_file.csv")}
      rawdata <- readr::read_csv("a_very_large_file.csv")
      `r ''````
  ````

- It's a good idea to regularly clear out all your caches with <span style="color: #008080;background-color:#9FDDBA">`knitr::clean_cache()`</span>.  

### Inline code

```{r}
comma <- function(x) format(x, digits = 2, big.mark = ',')
# `r comma(3452345)` to produce the inline result
```

This year has felt `r comma(3452345)` days long. Formatting decimals is a breeze using the `comma` function: `r comma(.12358124331)`.

### Exercises

1.  Set up a network of chunks where `d` depends on `c` and `b`, and
    both `b` and `c` depend on `a`. Have each chunk print `lubridate::now()`,
    set `cache = TRUE`, then verify your understanding of caching.
    
    ```{r cat3, echo = FALSE, comment = ""}
    cat(readr::read_file("docs/ex_for_ch21.Rmd"))
    ```
    
    <br> <br>
    
    [Rendered file](ex_for_ch21.html)

