---
title: "Chapter 22 - Graphics for communication"
author: "Vebash Naidoo"
date: "09/11/2020"
output: html_document
---

```{css, echo = FALSE}
.tabset h2 {display: none;}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE, message=FALSE,
                      comment = "#>", 
                      collapse = TRUE)

options(scipen=10000)
library(tidyverse)
library(flair)
library(magrittr)
library(lubridate)
library(werpals) # my own package - sciencificity github
```
# ggplot {#buttons .tabset .tabset-fade .tabset-pills}

__Click on the tab buttons below for each section__

<h2>Introduction</h2>

## Label

- You add labels with the <span style="color: #008080;background-color:#9FDDBA">
`labs()`</span>  function.

  * `title`: The purpose of a plot title is to summarise the main finding.
  * `subtitle`: adds additional detail in a smaller font beneath the title.
  * `caption`: adds text at the bottom right of the plot, often used to describe
    the source of the data.
  * `x`: replaces x-axis name (sometimes you want more descriptive names, or
  units).
  * `y`: replaces y-axis name (sometimes you want more descriptive names, or
  units).
  * The legend titles are replaced using the aesthetic that made it - e.g.
  `colour` or `fill`.
  
```{r plot1, include = FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(title = "Fuel efficiency generally decreases with engine size")
```

```{r, echo=FALSE}
decorate("plot1") %>% 
  flair("labs", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("title", background = "#cfdad4", 
        color = "#6d6875")
```

```{r plot2, include = FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(title = str_glue("Fuel efficiency generally decreases ",
                        "with engine size"),
       subtitle = str_glue("Two seaters (sports cars) are the ",
                           "exception because of their light weight"),
       caption = "Data from fueleconomy.gov")
```

```{r, echo=FALSE}
decorate("plot2") %>% 
  flair("subtitle", background = "#cfdad4", 
        color = "#6d6875") %>% 
  flair("caption", background = "#cfdad4", 
        color = "#6d6875")
```

```{r plot3, include = FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    colour = "Car type"
  )
```

```{r, echo=FALSE}
decorate("plot3") %>% 
  flair("x", background = "#cfdad4", 
        color = "#6d6875") %>% 
  flair("colour", background = "#cfdad4", 
        color = "#6d6875")
```


It's possible to use mathematical equations instead of text strings. 
Just switch `""` out for <span style="color: #008080;background-color:#9FDDBA">`quote()`</span>
and read about the available options in `?plotmath`.

```{r plot4, fig.asp = 1, out.width = "50%", fig.width = 3, include = FALSE}
df <- tibble(
  x = runif(10),
  y = runif(10)
)
ggplot(df, aes(x, y)) +
  geom_point() +
  labs(
    x = quote(sum(x[i]^2, i == 1, n)),
    y = quote(alpha + beta + frac(delta, theta))
  )
```

```{r, echo=FALSE}
decorate("plot4") %>% 
  flair("quote", background = "#9FDDBA", 
        color = "#008080")
```


### Exercises

1.  Create one plot on the fuel economy data with customised `title`,
    `subtitle`, `caption`, `x`, `y`, and `colour` labels.
    
    ```{r}
    ggplot(mpg, aes(class, hwy)) +
      geom_boxplot() +
      geom_jitter(aes(colour = displ), width = 0.1, height = 0) +
      labs(title = "Fuel efficiency is worst for pickups and suv",
           subtitle = "These vehicle types have the bigger engine size",
           x = "Car Type",
           y = "Highway fuel economy (mpg)",
           colour = "Engine size"
      )
    ```
    
1.  The `geom_smooth()` is somewhat misleading because the `hwy` for
    large engines is skewed upwards due to the inclusion of lightweight
    sports cars with big engines. Use your modelling tools to fit and display
    a better model.
    
    ```{r}
    ggplot(mpg, aes(displ, hwy, colour = class)) +
      geom_point() +
      geom_smooth(method = "lm", se = FALSE) +
      scale_colour_disney("pan") +
      labs(title = str_glue("Fuel efficiency generally decreases ",
                            "with engine size"),
           subtitle = str_glue("Two seaters (sports cars) are the ",
                               "exception because of their light weight"),
           caption = "Data from fueleconomy.gov",
           x = "Engine displacement (L)",
           y = "Highway fuel economy (mpg)",
           colour = "Car type")
    ```
    
1.  Take an exploratory graphic that you've created in the last month, and add
    informative titles to make it easier for others to understand.
    
<h2>Annotations</h2>
## Annotations

It's also often useful to label individual observations or groups of 
observations. The first tool you have at your disposal is 
<span style="color: #008080;background-color:#9FDDBA">`geom_text()`</span>. 
`geom_text()` is similar to `geom_point()`, but it has an additional aesthetic
- `label`. This makes it possible to add textual labels to your plots.

- a tibble that provides labels
- ?? find out in exercises

```{r annot1, include=FALSE}
mpg %>% 
  select(class, displ, hwy, model) %>% 
  group_by(class) %>% 
  mutate(rn = row_number(desc(hwy))) %>% 
  filter(rn %in% c(1,2))
         
# create a tibble for your label data
best_in_class <- mpg %>% 
  group_by(class) %>% 
  filter(row_number(desc(hwy)) == 1)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), # the label should be the model in your tbl 
            # use your label data
            data = best_in_class)
```

```{r, echo=FALSE}
decorate("annot1") %>% 
  flair("geom_text", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("label", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("model", background = "#cfdad4", 
        color = "#6d6875") %>% 
  flair("data = best_in_class", background = "#cfdad4", 
        color = "#6d6875")
```

It is sometimes hard to read, like here, where labels overlap with each other, 
and with the points. We can make things a little better by switching to 
<span style="color: #008080;background-color:#9FDDBA">`geom_label()`</span> 
which draws a rectangle behind the text. 
We also use the <span style="color: #008080;background-color:#9FDDBA">`nudge_y`</span>
parameter to move the labels slightly above the corresponding points.

```{r annot2, include=FALSE}
ggplot(data = mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_label(aes(label = model), 
            data = best_in_class,
            nudge_y = 2,
            alpha = 0.5)
```

```{r, echo=FALSE}
decorate("annot2") %>% 
  flair("geom_label", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("nudge_y", background = "#9FDDBA", 
        color = "#008080")  
```

There is still a problem in that there are two labels on top of each other in the
left corner. We can use the <span style="color: #008080;background-color:#9FDDBA">__ggrepel__</span> package by Kamil Slowikowski which automatically adjusts labels so that 
they don't overlap `r emo::ji("exclamation")` `r emo::ji("heart")`

```{r annot3, include=FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  # add a second layer of large, hollow points to 
  # highlight the points labelled
  geom_point(size = 3, shape = 1, 
             data = best_in_class)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_point(size = 3, shape = 1,
             data = best_in_class) +
  ggrepel::geom_label_repel(aes(label = model), 
                           data = best_in_class)
```

```{r, echo=FALSE}
decorate("annot3") %>% 
  flair("ggrepel::geom_label_repel", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("size = 3, shape = 1", background = "#9FDDBA", 
        color = "#008080")  
```

`theme(legend.position = "none"`) turns the legend off.

```{r}
# find the median displ, hwy for each class
class_avg <- mpg %>% 
  group_by(class) %>% 
  summarise(
    displ = median(displ),
    hwy = median(hwy)
  )

ggplot(mpg, aes(displ, hwy, 
                # move up here since used by both 
                # geom_label_repel - to colour label text
                # and geom_point to colour points
                colour = class)) +
  ggrepel::geom_label_repel(aes(label = class), # stick label in median place
                             data = class_avg,
                             size = 6,
                             label.size = 0,
                             segment.colour = NA) +
  geom_point() +
  # we've replaced our legend by labels on the plot
  # in the median position!
  theme(legend.position = "none")
```

You may want to add a single label to the plot, but you'll still need to 
create a data frame. <br>
Often, you want the label in the corner of the plot, so it's convenient to 
create a new data frame using `summarise()` to compute the __maximum values of x and y__.

```{r annot4, include=FALSE}
label <- mpg %>% 
  summarise(
    displ = max(displ),
    hwy = max(hwy),
    label = "Increasing engine size is \nrelated to decreasing fuel economy."
  )

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label),
            data = label, 
            vjust = "top", hjust = "right")
```

```{r, echo=FALSE}
decorate("annot4") %>% 
  flair("vjust", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("top", background = "#cfdad4", 
        color = "#6d6875") %>% 
  flair("hjust", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("right", background = "#cfdad4", 
        color = "#6d6875")
```

If you want to place the text exactly on the borders of the plot, you can 
use `+Inf` and `-Inf`. Note we use tibble here since we have more "static"
data.

```{r annot5, include = FALSE}
label <- tibble(
  displ = Inf,
  hwy = Inf,
  label = "Increasing engine size is \nrelated to decreasing fuel economy."
)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label), data = label, 
            vjust = "top", hjust = "right")
```

```{r, echo=FALSE}
decorate("annot5") %>% 
  flair("Inf", background = "#cfdad4", 
        color = "#6d6875")
```

We manually broke the label up into lines using `"\n"` but we may have
used `stringr::str_wrap()` to automatically add line breaks.

```{r}
"Increasing engine size is related to decreasing fuel economy." %>%
  # break the above text up automatically
  # I want 40 characters per line
  stringr::str_wrap(width = 40) %>%
  writeLines()
```

```{r annot6, include = FALSE}
label <- tibble(
  displ = Inf,
  hwy = Inf,
  label = stringr::str_wrap("Increasing engine size is 
                            related to decreasing fuel economy.", 
                            width = 35)
)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label), data = label, 
            vjust = "top", hjust = "right")
```

```{r, echo=FALSE}
decorate("annot6") %>% 
  flair("stringr::str_wrap", background = "#9FDDBA", 
        color = "#008080")
```

```{r just, echo = FALSE, fig.cap = "All nine combinations of `hjust` and `vjust`.", fig.asp = 0.5, fig.width = 4.5, out.width = "60%"}
vjust <- c(bottom = 0, center = 0.5, top = 1)
hjust <- c(left = 0, center = 0.5, right = 1)
df <- tidyr::crossing(hj = names(hjust), vj = names(vjust)) %>%
  mutate(
    y = vjust[vj],
    x = hjust[hj],
    label = paste0("hjust = '", hj, "'\n", "vjust = '", vj, "'")
  )
ggplot(df, aes(x, y)) +
  geom_point(colour = "grey70", size = 5) +
  geom_point(size = 0.5, colour = "red") +
  geom_text(aes(label = label, hjust = hj, vjust = vj), size = 4) +
  labs(x = NULL, y = NULL) 
```

### Exercises

1.  Use `geom_text()` with infinite positions to place text at the
    four corners of the plot.
    
    I am not sure if it is just me but I am unable to do it with
    Inf and -Inf. 
    
    So I used 0, 1 as the authors did for the `hjust`, `vjust` plot,
    and tweaked it for my case here.

    ```{r}
    vjust <- c("top", "top", "bottom", "bottom")
    hjust <- c("left", "right", "left", "right")
    label <- c("Top left", "Top right", "Bottom left", "Bottom right")
    x <- c(0, 1, 0, 1)
    y <- c(1, 1, 0, 0)
    df <- tibble(x = x,
                 y = y,
                 label = label,
                 vjust = vjust,
                 hjust = hjust)
    
    ggplot(df, aes(x, y)) +
      geom_text(aes(label = label),
                vjust = vjust,
                hjust = hjust) +
      scale_x_continuous(labels = NULL) +
      scale_y_continuous(labels = NULL) +
      labs(x = "",
           y = "")
    ```
    

1.  Read the documentation for `annotate()`. How can you use it to add a text
    label to a plot without having to create a tibble?
    
    Some examples from the help page below. Used when you want to add some
    static text. I like the use of it for things like $italic(R) ^ 2 == 0.75$,
    and also seems useful for marking anomalies on a plot.
    
    ```{r}
    p <- ggplot(mtcars, aes(x = wt, y = mpg)) + geom_point()
    p + annotate("text", x = 2:5, y = 25, label = "Some text")
    p + annotate("rect", xmin = 3, xmax = 4.2, ymin = 12, ymax = 21,
      alpha = .2)
    p + annotate("text", x = 2:3, y = 20:21, label = c("my label", "label 2"))
    p + annotate("text", x = 4, y = 25, label = "italic(R) ^ 2 == 0.75",
      parse = TRUE)
    ```
    

1.  How do labels with `geom_text()` interact with faceting? How can you
    add a label to a single facet? How can you put a different label in
    each facet? (Hint: think about the underlying data.)
    
    ```{r}
    # Got this example from this SO post
    # https://stackoverflow.com/questions/15867263/ggplot2-geom-text-with-facet-grid
    x <-runif(9, 0, 125) 
    data <- as.data.frame(x) 
    data$y <- runif(9, 0, 125) 
    data$yy <- factor(c("a","b","c")) 
    
    ggplot(data, aes(x, y)) + 
        geom_point(shape = 2) + 
        facet_grid(~yy) + 
        geom_text(aes(x, y, label=lab),
            data=data.frame(x=60, y=Inf, lab=c("this","is","the way"),
                 yy=letters[1:3]), vjust=1)
    
    ggplot(data, aes(x, y)) + 
        geom_point(shape = 2) + 
        facet_grid(~yy) + 
        geom_text(aes(x, y, label=lab),
            # text on one facet
            data=data.frame(x=60, y=Inf, lab=c("this is the only way"),
                 yy=letters[1]), # one facet only
            vjust=1)
    ```

1.  What arguments to `geom_label()` control the appearance of the background
    box?
    
    - label.padding: Amount of padding around label. Defaults to 0.25 lines.
    - label.r: Radius of rounded corners. Defaults to 0.15 lines.
    - label.size: Size of label border, in mm.

1.  What are the four arguments to `arrow()`? How do they work? Create a series
    of plots that demonstrate the most important options.
    
    Tweaked from this [Stack Overflow post](https://stackoverflow.com/questions/38008863/how-to-draw-a-nice-arrow-in-ggplot2).
    
    ```{r}
    ggplot(mtcars, aes(wt, mpg)) + 
      geom_point() +
      geom_segment(aes(x = 5, y = 30, xend = 3.5, yend = 25),
                      arrow = arrow(length = unit(0.5, "cm")))
    
    ggplot(mtcars, aes(wt, mpg)) + 
      geom_point() +
      geom_segment(aes(x = 5, y = 30, xend = 3.5, yend = 25),
                      arrow = arrow(length = unit(0.5, "cm"),
                                    type = "closed"))
    
    ggplot(mtcars, aes(wt, mpg)) + 
      geom_point() +
      geom_segment(aes(x = 5, y = 30, xend = 3.5, yend = 25),
                      arrow = arrow(length = unit(0.5, "cm"),
                                    type = "closed",
                                    ends = "both"))
    
    ggplot(mtcars, aes(wt, mpg)) + 
      geom_point() +
      geom_segment(aes(x = 5, y = 30, xend = 3.5, yend = 25),
                      arrow = arrow(angle = 60, # how narrow or wide the arrow head
                                    length = unit(0.5, "cm"),
                                    type = "closed",
                                    ends = "both"))
    
    
    ggplot(mtcars, aes(wt, mpg)) + 
      geom_point() +
      geom_segment(aes(x = 5, y = 30, xend = 3.5, yend = 25),
                      arrow = arrow(angle = 15, # how narrow or wide the arrow head
                                    length = unit(0.5, "cm"),
                                    type = "closed",
                                    ends = "both"))
    ```

<h2>Scales</h2> 
## Scales

You can also make your plot better for communication by adjusting the scales. 
ggplot2 automatically adds sensible default scales behind the scenes for you.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))
```

Note this is the same as above.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete()
```

### Axis ticks and legend keys

Collectively axes and legends are called <span style="color: #008080;background-color:#9FDDBA">__guides__</span>. 
Axes are used for x and y aesthetics; legends are used for everything else.

- <span style="color: #008080;background-color:#9FDDBA">`breaks`</span> 
controls the position of the axis ticks, or the values associated 
with the legend keys. 
- <span style="color: #008080;background-color:#9FDDBA">`labels`</span> controls 
the text label associated with each tick/key. Supply 
a character vector the same length as `breaks` OR suppress labels by using
<span style="color: #008080;background-color:#9FDDBA">`NULL`</span> .
- When is `NULL` appropriate? Useful for maps, or when you can't share
absolute numbers in a published plot.
- Breaks and labels for datetimes is a bit different: 

  * `date_labels` takes a format specification, in the same form as
    `parse_datetime()`.
  
  * `date_breaks`, takes a string like "2 days" or "1 month".

```{r br1, include = FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_y_continuous(breaks = seq(15, 40, by = 5))
```

```{r, echo=FALSE}
decorate("br1") %>% 
  flair("scale_y_continuous", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("breaks", background = "#cfdad4", 
        color = "#6d6875")
```

```{r br2, include = FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL)
```

```{r, echo=FALSE}
decorate("br2") %>% 
  flair("labels = NULL", background = "#cfdad4", 
        color = "#6d6875")
```

```{r}
presidential %>% 
  # add the number of the president, 
  # e.g. Obama was the 44th president of USA
  mutate(id = 33 + row_number())

presidential %>% 
  # add the number of the president, 
  # e.g. Obama was the 44th president of USA
  mutate(id = 33 + row_number()) %>%  
  ggplot(aes(start, id)) +
  geom_point() +
  geom_segment(aes(xend = end, yend = id)) +
  scale_x_date(NULL,
               breaks = presidential$start,
               date_labels = "'%y") +
  ggrepel::geom_label_repel(aes(label = name),
                                nudge_x = 0.5,
                                nudge_y = 0.4)
```


### Legend layout

To control the overall position of the legend, you need to use a <span style="color: #008080;background-color:#9FDDBA">`theme()`</span>  
setting. The theme setting `legend.position` controls where the legend is drawn.
You can also use `legend.position = "none"` to hide the display of the legend.

```{r fig.asp = 1, fig.align = "default", out.width = "75%", fig.width = 4}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))
base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") # the default
```

- To control the display of individual legends, use `guides()` along with `guide_legend()` or `guide_colourbar()`. 
- Control the number of rows the legend uses with `nrow`
- Override one of the aesthetics to make the points bigger. Useful if you have 
used a low `alpha` to display many points on a plot.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 1,
                               override.aes = list(size = 4)))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 4, 
                               override.aes = list(size = 5)))
```

### Replacing a scale

Sometimes you will want to replace a scale completely - we have done
this before when we log transformed graphs.

__Aesthetic mapping Transform__

```{r, fig.align = "default"}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d()
# disadvantage of this transformation is that the axes are 
# now labelled with the transformed values, making it hard to interpret the plot
ggplot(diamonds, aes(log10(carat), log10(price))) +
  geom_bin2d()
```

__Scale Transform__

Visually identical to above, except the axes are labelled using the original 
data scale.

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()
```

Useful colour scale alternatives are the ColorBrewer scales for certain colour 
blindness - e.g. red-green colour blindness.

```{r, fig.align = "default"}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv))

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1")
```

Adding a redundant shape mapping is also completely acceptable. It also
ensures the plot is interpretable in black and white.

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) +
  scale_colour_brewer(palette = "Set1")
```

In ColorBrewer the sequential (top) and diverging (bottom) are useful if your 
categorical values are __ordered__, or have a "middle". Often the case if you've 
used `cut()` to make a continuous variable into a categorical variable.

```{r brewer, fig.asp = 2.5, echo = FALSE, fig.cap = "All ColourBrewer scales."}
par(mar = c(0, 3, 0, 0))
RColorBrewer::display.brewer.all()
```

Use `scale_colour_manual()` when you have a predefined mapping between values and colours. 

```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue"))
```

- For continuous colour, use `scale_colour_gradient()` or `scale_fill_gradient()`. 
- For a diverging scale, use `scale_colour_gradient2()`.  Here positive and 
negative values may have different colours. Also useful if you want to 
distinguish points above or below the mean.
- `scale_colour_viridis()` is also a good alternative.

```{r, fig.align = "default", fig.asp = 1, fig.width = 4}
df <- tibble(
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() +
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```

All colour scales have `scale_colour_x()` and `scale_fill_x()` for the `colour` 
and `fill` aesthetics.

### Exercises

1.  Why doesn't the following code override the default scale?

    The hex should use __fill__ not __colour__

    ```{r}
    ggplot(df, aes(x, y)) +
      geom_hex() +
      scale_colour_gradient(low = "white", high = "red") +
      coord_fixed()
    ```
    
    ```{r}
    ggplot(df, aes(x, y)) +
      geom_hex() +
      scale_fill_gradient(low = "white", high = "red") +
      coord_fixed()
    ```

1.  What is the first argument to every scale? How does it compare to `labs()`?
    
    The `...` argument which 	passes on to the particular
    scale function in the arguments. The first arg is aesthetics that this 
    scale works with e.g. `x`, `y`, `colour` etc.
    
    `labs` is similar but talks to the labels not the scale.
    

1.  Change the display of the presidential terms by:

    a. Combining the two variants shown above.
    
      ```{r}
      presidential %>%
        mutate(id = 33 + row_number()) %>%
        ggplot(aes(start, id, colour = party)) +
          geom_point() +
          geom_segment(aes(xend = end, yend = id)) +
          scale_colour_manual(values = c(Republican = "red", 
                                         Democratic = "blue")) +
          scale_x_date(NULL,
                     breaks = presidential$start,
                     date_labels = "'%y")
      ```
    
    a. Improving the display of the y axis.
    
        ```{r}
        presidential_amend <- presidential %>%
          mutate(id = 33 + row_number(),
                 id_label = str_c(id, c(rep("th ", 7), 
                                  "st ", "nd ", "rd ", "th "), 
                                  c(rep("President", 11))),
                 id_label = fct_reorder(id_label, id)) 
        
        presidential_amend %>%
          ggplot(aes(start, id, colour = party)) +
            geom_point() +
            geom_segment(aes(xend = end, yend = id)) +
            scale_colour_manual(values = c(Republican = "red", 
                                           Democratic = "blue")) +
            scale_x_date(NULL,
                       breaks = presidential_amend$start,
                       date_labels = "'%y") +
            scale_y_continuous(breaks = presidential_amend$id,
                               labels = presidential_amend$id_label) +
            ggrepel::geom_label_repel(aes(label = name),
                                    nudge_x = 0.5,
                                    nudge_y = 0.4) +
            labs(y = "",
                 x = "Presidential term",
                 title = str_wrap("Republicans
                 seem to be winning the race in terms of time in office?", 
                 width = 70))
        ```    
    a. Labelling each term with the name of the president.
    
        Done above.
        
    a. Adding informative plot labels.
    
        Done above.
        
    a. Placing breaks every 4 years (this is trickier than it seems!).
    
        ```{r}
            presidential_amend <- presidential %>%
              mutate(id = 33 + row_number(),
                     id_label = str_c(id, c(rep("th ", 7), 
                                      "st ", "nd ", "rd ", "th "), 
                                      c(rep("President", 11))),
                     id_label = fct_reorder(id_label, id)) 
        
            breaks <- function(df){
              df %>% 
                mutate(yr_start = year(start),
                       yr_end = year(end)) %>% 
                summarise(min_start = min(yr_start),
                       max_end = max(yr_end))
            }
            
            break_df <- presidential_amend %>% 
             breaks() 
            
            presidential_amend %>%
              ggplot(aes(start, id, colour = party)) +
                geom_point() +
                geom_segment(aes(xend = end, yend = id)) +
                scale_colour_manual(values = c(Republican = "red", 
                                               Democratic = "blue")) +
                scale_x_date(NULL,
                             breaks = seq.Date(ymd(str_glue("{break_df$min_start}0101")), 
                                          ymd(str_glue("{break_df$max_end}0101")), 
                                          "4 years"),
                             date_labels = "'%y") +
                scale_y_continuous(breaks = presidential_amend$id,
                                   labels = presidential_amend$id_label) +
                ggrepel::geom_label_repel(aes(label = name),
                                        nudge_x = 0.5,
                                        nudge_y = 0.4) +
                labs(y = "",
                     x = "Presidential term",
                     title = str_wrap("Republicans
                     seem to be winning the race in terms of time in office?", 
                     width = 70))
            ``` 
1.  Use `override.aes` to make the legend on the following plot easier to see.

    ```{r, dev = "png", out.width = "80%"}
    ggplot(diamonds, aes(carat, price)) +
      geom_point(aes(colour = cut), alpha = 1/20)
    ```
    
    ```{r, dev = "png", out.width = "80%"}
    ggplot(diamonds, aes(carat, price)) +
      geom_point(aes(colour = cut), alpha = 1/20) +
      guides(colour = guide_legend(override.aes = list(alpha = 1)))
    ```

## Zooming

You can control plot limits by:

1. Adjusting what data are plotted
1. Setting the limits in each scale
1. Setting `xlim` and `ylim` in `coord_cartesian()`

To zoom in on a region of the plot, use `coord_cartesian()`. 

```{r fig.align = "default", message = FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth() +
  # zoom in to a specific set of coords
  coord_cartesian(xlim = c(5,7), ylim = c(10,30))

mpg %>% 
  filter(displ >= 5 & displ <= 7,
         hwy >= 10 & hwy <= 30) %>% 
  ggplot(aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  # we draw a smoothing fn for the data
  # but we filtered obs to just the ones we want to see
  geom_smooth()
```

Notice above that `coord_cartesian()` is the true zoom in, the smoothing function
considers all data. In the filtered graph the smoothing function considers
only the filtered observations so it is a very different view of the data.
Consider your use case for zooming then choose a method.

```{r fig.align = "default", fig.width = 4}
suv <- mpg %>% filter(class == "suv")
compact <- mpg %>% filter(class == "compact")
ggplot(suv, aes(displ, hwy, colour = drv)) +
  geom_point()
ggplot(compact, aes(displ, hwy, colour = drv)) +
  geom_point()
```

The scales above are however different so comparisons are not easy. 
You may share scales across multiple plots.

```{r out.width = "50%", fig.align = "default", fig.width = 4}
x_scale <- scale_x_continuous(limits = range(mpg$displ))
y_scale <- scale_y_continuous(limits = range(mpg$hwy))
col_scale <- scale_colour_discrete(limits = unique(mpg$drv))
ggplot(suv, aes(displ, hwy, colour = drv)) +
  geom_point() +
  x_scale +
  y_scale +
  col_scale
ggplot(compact, aes(displ, hwy, colour = drv)) +
  geom_point() +
  x_scale +
  y_scale +
  col_scale
```

We could have used faceting but this technique is useful if for instance, 
you want spread plots over multiple pages of a report.

<h2>Themes</h2>
## Themes

Non-data elements of your plot can be amended with a theme.

```{r, message = FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw()

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  theme_light()

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  theme_dark()
```

