---
title: "Chapter 23 - Rmd Formats"
author: "Vebash Naidoo"
date: "26/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set output format

There are two ways to set the output of a document:

1.  Modify the YAML header: 
    
    ```yaml
    title: "Viridis Demo"
    output: html_document
    ```
    
1.  Call `rmarkdown::render()` (useful if you want to produce multiple types of output):
    
    ```{r eval = FALSE}
    rmarkdown::render("diamond-sizes.Rmd", output_format = "word_document")
    ```
    
To get help: `?rmarkdown::html_document`    

## Useful options

1. Sending to decision makers? Use <span style="color: #008080;background-color:#9FDDBA">echo = FALSE</span> in opts_chunk.

    ```{r, eval = FALSE}
    knitr::opts_chunk$set(echo = FALSE)
    ```
    
1. Want to include code but hide it? Use <span style="color: #008080;background-color:#9FDDBA">code_folding</span> in YAML Header.

    ```yaml
    output:
      html_document:
        code_folding: hide
    ```

## Dashboard

You can create a flexdashboard using the code:

```{r comment = "", echo = FALSE}
cat(readr::read_file("docs/flex_trial.Rmd"))
```

The rendered dashboard is [here](flex_trial.html).

## Interactivity

```{r}
library(leaflet)
leaflet() %>%
  setView(174.764, -36.877, zoom = 16) %>% 
  addTiles() %>%
  addMarkers(174.764, -36.877, popup = "Maungawhau") 
```
