---
title: "Chapter 2 - Workflow basics"
author: "Vebash Naidoo"
date: "23/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(tidyverse)
library(flair)
```

## Workflow ...

<span style="color: #008080;background-color:#9FDDBA">`object_name <- value`</span> can be read as "object name __gets__ value"

Find the `<-` tedious to enter? RStudio has a shortcut `Alt + -` to enter it for you.

<image src = assets/alt_enter.gif>

Alternately there's the great {styler} package that you may use to convert your code to the preferred style.

1. Go to Tools -> Install Packages and install `styler`
1. Else go to the [Github page](https://github.com/r-lib/styler) to install.
1. In your code select the section you want to format (not necessary if you want to do for the entire file).

    * Go to Addins
    * STYLER
    * Choose `Style selection` or the option appropriate for your task.
  
<image src = assets/styler.gif>  

Sometime you have an assignment, but you also want to print your variable to view its contents. Do this by surrounding the <span style="color: #008080;background-color:#9FDDBA">assignment</span> with <span style="color: #008080;background-color:#9FDDBA">()</span>.

```{r save_and_out, include = FALSE}
( y <- seq(1, 10, length.out = 5) ) 
```
```{r, echo = FALSE}
decorate("save_and_out") %>% 
  flair("( ",
        background = "#9FDDBA", 
        color = "#008080") %>% 
  flair(" )",
        background = "#9FDDBA", 
        color = "#008080")
```

### Exercises

1.  Why does this code not work?

    ```{r, error = TRUE}
    my_variable <- 10
    my_varıable
    ```
    
    Look carefully! (This may seem like an exercise in pointlessness, but
    training your brain to notice even the tiniest difference will pay off
    when programming.)
    
    It doesn't work because the `i` in the `my_variable` has been replaced by another character `ı`.
    
1.  Tweak each of the following R commands so that they run correctly:

    ```{r, eval = FALSE}
    library(tidyverse)
    ggplot(dota = mpg) + 
      geom_point(mapping = aes(x = displ, y = hwy))
    
    fliter(mpg, cyl = 8)
    filter(diamond, carat > 3)
    ```    
    
  ```{r ex2_1, include = FALSE}
  library(tidyverse)
  ggplot(data = mpg) +
    geom_point(mapping = aes(x = displ, y = hwy))
  ```
    ```{r ex2_2, include = FALSE}
    filter(mpg, cyl == 8)
    ```
    ```{r ex2_3, include = FALSE}
    filter(diamonds, carat > 3)
    ```
    
    ```{r, echo=FALSE}
    decorate("ex2_1") %>% 
      flair("data", 
            background = "#9FDDBA", 
            color = "#008080")
    ```
    ```{r, echo=FALSE}
    decorate("ex2_2") %>% 
      flair("==", 
            background = "#9FDDBA", 
            color = "#008080") %>%       
      flair("filter(",
            background = "#9FDDBA", 
            color = "#008080")
    ```    
    ```{r, echo=FALSE}
    decorate("ex2_3") %>% 
      flair("diamonds",
            background = "#9FDDBA", 
            color = "#008080")      
    ```        
1.  Press Alt + Shift + K. What happens? How can you get to the same place
    using the menus?
    
    It opens the shortcuts popup page. You may _close_ the page 
    by pressing __`ESC`__. 
    
    To get there through the menus go to Tools -> Keyboard Shortcuts 
    -> Help