---
title: "Chapter 7 - Tibbles"
author: "Vebash Naidoo"
date: "17/10/2020"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
options(scipen=10000)
library(tidyverse)
library(flair)
library(nycflights13)
library(palmerpenguins)
library(gt)
library(skimr)
library(emo)
library(tidyquant)
library(lubridate)
library(magrittr)
theme_set(theme_tq())
```

## Class 

```{r tib1, include=FALSE}
class(iris) # initially R's data.frame
(iris2 <- as_tibble(iris))
class(iris2) # "tbl_df"     "tbl"        "data.frame"
```

```{r, echo=FALSE}
decorate("tib1") %>% 
  flair("as_tibble", background = "#9FDDBA", 
        color = "#008080")
```

The <span style="color: #008080;background-color:#9FDDBA">`tbl`</span> and 
<span style="color: #008080;background-color:#9FDDBA">`tbl_df`</span> 
tell us this has been converted to a tibble.


## Creation

To create a `tibble` we use the <span style="color: #008080;background-color:#9FDDBA">`tibble()`</span> function.

```{r tib2, include=FALSE}
tibble(
  x = 1:5,
  y = 1, # recycled
  z = x ^ 2 + y
)
```

```{r, echo=FALSE}
decorate("tib2") %>% 
  flair("tibble", background = "#9FDDBA", 
        color = "#008080")
```

A tibble is just a different form of a data.frame. The column names
do not need to comply with R's variable naming convention.

For example the variable name `test subject` is unacceptable in R as a var name,
however it is totally fine if it is a column name in a tibble.
The variable must be referred to by using the backticks `` or "".

```{r tib3, include=FALSE}
# Ch7 in R4DS - tibbles
# Can use either backticks `` or quotation marks "" to surround non-syntactic names
(tb = tibble(
  ":)" = "smile",
  " " = "space",
  `2000` = "number",
  `some non syntactic name` = "some value",
  meaning = 42 # no back ticks or "" since this is syntactically allowed
))

# Can use `` or "" to access the variable with the non-syntactic name
tb$`:)`
tb$":)"

# Can use `` or "" to manipulate the non-syntactic named var
tb %>% 
  mutate(":)" = "smiley")

tb %>% 
  mutate(`:)` = "smiley too")
```

```{r, echo=FALSE}
decorate("tib3") %>% 
  flair("tibble", background = "#9FDDBA", 
        color = "#008080")
```

The variable `:)` in tb contains the value: 
<span style="color: #008080;background-color:#9FDDBA">`r tb$":)"`</span>. 
The variable `some non syntactic name` in tb contains the value: 
<span style="color: #008080;background-color:#9FDDBA">
`r tb$"some non syntactic name"`</span>.

<!--
To output non-syntactic variable names in Rmd you use: 
`r tb$"some non syntactic name"`
SO Post: https://stackoverflow.com/questions/24472114/use-a-variable-name-with-spaces-inline-in-r-markdown
-->

Another way is to create a tibble is to use the <span style="color: #008080;background-color:#9FDDBA">`tribble()`</span> function with `~` used 
for each column name. I often use the datapasta `r emo::ji("package")` to paste
a tribble from something I have copied onto the clipboard.
<br>

<img src= assets/datapasta_paste_tribble.gif alt="pasting a tribble from data on clipboard" width="850" height="400">

<br>


A `tribble` is _transposed tibble_. It makes it easy to lay out small amounts
of data in an easy to read format.


```{r tib4, include=FALSE}
tribble(
  ~x, ~y, ~z,
  #--/--/----
  "a", 2, 3.6,
  "b", 1, 8.5
)

tibble::tribble(
                               ~Flavour, ~Total.2009, ~Total.2011, ~East, ~Midwest, ~South, ~West, ~Rep, ~Dem, ~Ind,
                            "Chocolate",         27L,         28L,   31L,      32L,    28L,   21L,  32L,  23L,  30L,
                              "Vanilla",         22L,         26L,   27L,      22L,    30L,   22L,  28L,  26L,  22L,
      "Cookie Dough/ Cookies and cream",         22L,         22L,   26L,      22L,    21L,   19L,  24L,  18L,  24L,
           "Butter Pecan/ Swiss Almond",         20L,         19L,   12L,      24L,    21L,   15L,  15L,  22L,  19L,
                  "Mint Chocolate Chip",         17L,         15L,   15L,      15L,    15L,   15L,  16L,  12L,  16L,
                           "Strawberry",         13L,         12L,    8L,      10L,    15L,   12L,  10L,  13L,  12L,
                           "Rocky Road",         14L,         11L,    8L,      11L,     8L,   19L,  14L,  11L,   9L,
                               "Coffee",          9L,          9L,   10L,       7L,     6L,   14L,   7L,  11L,   8L,
                        "Peanut Butter",          8L,          8L,   10L,       9L,     7L,    8L,  12L,   5L,   8L,
                       "Cherry Vanilla",          9L,          7L,   10L,       6L,     7L,    7L,   6L,  10L,   8L,
                            "Pistachio",          8L,          7L,    7L,       6L,     6L,    8L,   5L,   9L,   6L,
                      "Black Raspberry",          6L,          6L,   10L,       6L,     3L,    6L,   7L,   5L,   6L,
                                "Peach",          4L,          5L,    6L,       4L,     7L,    3L,   5L,   7L,   4L,
  "Seasonal, such as pumpkin or eggnog",          2L,          4L,    4L,       4L,     5L,    4L,   5L,   3L,   5L,
                                "Other",          9L,         13L,   13L,      12L,    13L,   15L,   9L,  16L,  13L,
                 "Do not eat ice cream",          3L,          3L,    1L,       3L,     2L,    5L,   2L,   3L,   3L
  )

```

```{r, echo=FALSE}
decorate("tib4") %>% 
  flair("tribble", background = "#9FDDBA", 
        color = "#008080")
```

## Printing

```{r}
tibble(
  a = lubridate::now() + runif(1e3) * 86400,
  b = lubridate::today() + runif(1e3) * 30,
  c = 1:1e3,
  d = runif(1e3),
  e = sample(letters, 1e3, replace = TRUE)
) %>% 
  print(n = 15, width = Inf) # show 15 rows instead of 10, and all cols
```

## Subsetting

```{r sub1, include=FALSE}
(df <- tibble(
  x = runif(5),
  y = rnorm(5)
))

# extract by name
df$x
# OR
df[["x"]]

# extract by position
df[[1]]

# When using a pipe use the placeholder . to access
df %>% 
  .$x

df %>% 
  .[["x"]]
```

```{r, echo=FALSE}
decorate("sub1") %>% 
  flair("$", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("[[", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("[.$]+", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("[.[[]+", background = "#9FDDBA", 
        color = "#008080")
```

## Summary

Tibbles:

- Print well, so as to not overwhelm your console (by printing 10
rows and only as many columns as will fit in your console window).

  * Want to print all columns? Use `print(n = xx, width = Inf)`
  <br>
- Subset more strictly that data.frame

  * Never partial matching
  * Warns if column you're trying to access does not exist.
  <br>
- Older functions may not work with tibbles. To convert a
tibble to a data.fram use <span style="color: #008080;background-color:#9FDDBA">`as.data.frame(tibble_name)`</span>
  

### Exercises

1.  How can you tell if an object is a tibble? (Hint: try printing `mtcars`,
    which is a regular data frame). 
    
    ```{r}
    mtcars
    class(mtcars)
    ```
    
    Tibbles:
    
    - Print only 10 rows, and as many columns as will fit on
    the console / screen
    - If you find the calss of a tibble among the output should
    be `tbl_df` and `tbl`.
    
    ```{r}
    penguins
    class(penguins)
    ```
    
1.  Compare and contrast the following operations on a `data.frame` and 
    equivalent tibble. What is different? Why might the default data frame
    behaviours cause you frustration?
    
    - On a data.frame
    
      ```{r}
      df <- data.frame(abc = 1, xyz = "a")
      df$x
      df[, "xyz"]
      df[, c("abc", "xyz")]
      ```
    
    - On a tibble
    
      ```{r, warning = TRUE}
      df <- tibble(abc = 1, xyz = "a")
      df$x
      df[, "xyz"]
      df[, c("abc", "xyz")]
      ```    
    
      * The data frame does not have an x variable, yet with a 
        data.frame it prints a value for x since it does partial
        matching.
      * The data frame has a character attribute for the column
        `xyz` yet the data.frame converts it to a factor, and 
        outputs the factors levels. `r emo::ji("oops")`
      * The information about the type of data contained in 
      every variable when you print a tibble is useful 
      `r emo::ji("perfect")`.

1.  If you have the name of a variable stored in an object, e.g. `var <- "mpg"`,
    how can you extract the reference variable from a tibble?
    
    ```{r}
    var <- "mpg"
    (test_tbl <- as_tibble(mtcars))
    
    # does not work: 
    # test_tbl$var
    # NULL 
    # Warning message:
    # Unknown or uninitialised column: `var`. 
    
    # does not work
    # test_tbl[["var"]]
    # NULL
    
    # Need to use:
    test_tbl[[var]]
    ```
    

1.  Practice referring to non-syntactic names in the following data frame by:

    1.  Extracting the variable called `1`.

    1.  Plotting a scatterplot of `1` vs `2`.

    1.  Creating a new column called `3` which is `2` divided by `1`.
        
    1.  Renaming the columns to `one`, `two` and `three`. 
    
    ```{r}
    annoying <- tibble(
      `1` = 1:10,
      `2` = `1` * 2 + rnorm(length(`1`))
    )
    ```
    
    ```{r}
    # Extracting the variable called `1`.
    annoying$`1`
    annoying$"1"
    annoying[["1"]]
    annoying[[1]] # by position
    # does not work
    # annoying[[`1`]]
    # Error in tbl_subset2(x, j = i, j_arg = substitute(i)) : 
    #   object '1' not found
    ```
    ```{r, warning = TRUE}
    # Plotting a scatterplot of `1` vs `2`.
    annoying %>% 
      ggplot(aes(x = `1`, y = `2`)) +
      geom_point()
    
    # discouraged ways
    annoying %>% 
      ggplot(aes(x = .$`1`, y = .$`2`)) +
      geom_point()
    
    annoying %>% 
      ggplot(aes(x = .$"1", y = .$"2")) +
      geom_point()
    
    annoying %>% 
      ggplot(aes(x = .[[1]], y = .[[2]])) +
      geom_point()
    ```
    
    ```{r}
    # Creating a new column called `3` which is 
    # `2` divided by `1`.
    (annoying <- annoying %>% 
      mutate(`3` = `2`/`1`))
    ```
    
    ```{r}
    # Renaming the columns to `one`, `two` and `three`.
    (annoying <- annoying %>% 
      rename("one" = "1",
             "two" = "2",
             "three" = "3"))
    ```

1.  What does `tibble::enframe()` do? When might you use it?

    It converts named vectors into tibbles.
    
    ```{r}
    # examples from the ?enframe help page
    enframe(1:3)
    enframe(c(a = 5, b = 7))
    enframe(list(one = 1, two = 2:3, three = 4:6))
    ```
    Note: Selecting examples on a help page in RStudio and pressing
    Ctrl + Enter sends the example to your console and runs it!
    
    <img src = assets/ctrl_enter_example_help_file2.gif 
    alt="Ctrl and Enter in Help File" 
    width="800" height="450">
    
1.  What option controls how many additional column names are printed
    at the footer of a tibble?  
    
    ```
    n_extra: Number of extra columns to print abbreviated information for, 
    if the width is too small for the entire tibble. If NULL, the default, 
    will print information about at most tibble.max_extra_cols extra columns.
    ```











