---
title: "Chapter 8 - Data Import"
author: "Vebash Naidoo"
date: "18/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
options(scipen=10000)
library(tidyverse)
library(flair)
library(nycflights13)
library(palmerpenguins)
library(gt)
library(skimr)
library(emo)
library(tidyquant)
library(lubridate)
library(magrittr)
theme_set(theme_tq())
```

__Inline csv file__

```{r}
read_csv("a,b,c
1,2,3
4,5,6")
```

__Skip some columns__

- metadata
- commented lines that you don't want to read

```{r}
read_csv("The first line of metadata
  The second line of metadata
  x,y,z
  1,2,3", skip = 2)

read_csv("# A comment I want to skip
  x,y,z
  1,2,3", comment = "#")
```

__No column names in data__

```{r}
read_csv("1,2,3\n4,5,6", # \n adds a new line 
         col_names = FALSE) # cols will be labelled seq from X1 .. Xn

read_csv("1,2,3\n4,5,6", 
         col_names = c("x", "y", "z")) # cols named as you provided here
```

__NA values__

```{r}
read_csv("a,b,c,d\nnull,1,2,.", 
         na = c(".",
                "null"))
# here we specify that the . and null
# must be considered to be missing values
```

### Exercises

1.  What function would you use to read a file where fields were separated with  
    "|"?
    
    `read_delim()`
    
    ```{r}
    # from the ?read_delim help page
    read_delim("a|b\n1.0|2.0", delim = "|")
    ```
    
    
1.  Apart from `file`, `skip`, and `comment`, what other arguments do
    `read_csv()` and `read_tsv()` have in common?
    
    All columns are common across the functions.
    
    - col_names
    - col_types
    - locale
    - na
    - quoted_na
    - quote
    - trim_ws
    - n_max 
    - guess_max 
    - progress 
    - skip_empty_rows 
    
    
1.  What are the most important arguments to `read_fwf()`?

    - file to read
    - col_positions as created by fwf_empty(), fwf_widths() or fwf_positions()
    which tells the function where a column starts and ends.
   
1.  Sometimes strings in a CSV file contain commas. To prevent them from
    causing problems they need to be surrounded by a quoting character, like
    `"` or `'`. By default, `read_csv()` assumes that the quoting
    character will be `"`. What argument to `read_csv()` do you need to specify
    to read the following text into a data frame?
    
    ```{r, eval = FALSE}
    "x,y\n1,'a,b'"
    ```
    
    Specify the quote argument.
    
    ```{r}
    read_csv("x,y\n1,'a,b'", quote = "'")
    ```
    
    
1.  Identify what is wrong with each of the following inline CSV files. 
    What happens when you run the code?
    
    ```{r, eval = FALSE}
    read_csv("a,b\n1,2,3\n4,5,6") # only 2 cols specified but 
    read_csv("a,b,c\n1,2\n1,2,3,4")
    read_csv("a,b\n\"1")
    read_csv("a,b\n1,2\na,b")
    read_csv("a;b\n1;3")
    ```
    
    > read_csv("a,b\n1,2,3\n4,5,6") <br>
    > only 2 cols specified but 3 values provided
    <br>
    
    > read_csv("a,b,c\n1,2\n1,2,3,4") <br>
    > 3 col names provided, but either too few, or too many column values provided
    <br>
    
    > read_csv("a,b\n\"1") <br>
    > 2 col names provided, but only one value provided. <br>
    > closing " missing
    <br>
    
    > read_csv("a,b\n1,2\na,b")
    > Nothing syntactically a problem, but the rows are filled<br>
    > with the column headings?
    <br>
    
    > read_csv("a;b\n1;3")
    > the read_csv2 which reads ; as delimiters should have been used
    <br>
    
    They all run, but most have warnings, and some are not imported as expected.
    
    ```{r, warning = TRUE}
    read_csv("a,b\n1,2,3\n4,5,6") # only 2 cols specified but 
    read_csv("a,b,c\n1,2\n1,2,3,4")
    read_csv("a,b\n\"1")
    read_csv("a,b\n1,2\na,b")
    read_csv("a;b\n1;3")
    ```
    
## Parsing

```{r import1a, include = FALSE}
str(parse_logical(c("TRUE", "FALSE", "NA")))
str(parse_integer(c("1", "2", "3")))
str(parse_date(c("2010-01-01", "1979-10-14")))
```

```{r, echo=FALSE}
decorate("import1a") %>% 
  flair("parse_", background = "#9FDDBA", 
        color = "#008080")
```

All <span style="color: #008080;background-color:#9FDDBA">`parse_xxx()`</span>
variants provide a uniform specification to use.

> parse_x(character_vector_to_parse,
> na = c(x, y))

```{r, warning = TRUE}
parse_integer(c("1", "231", ".", "456"), na = ".")
(x <- parse_integer(c("123", "345", "abc", "123.45")))
```

To detect problems use `problems()`.

```{r import2, include=FALSE}
problems(x)
```

```{r, echo=FALSE}
decorate("import2") %>% 
  flair("problems", background = "#9FDDBA", 
        color = "#008080")
```

Sometimes depending on where in the world you are you will have different
conventions when it comes to numbers.

For example you may separate the integer part from the decimal part by
using a `.` or a `,`. To tell the parsing function what kind of data
you're expecting to be in a vector use <span style="color: #008080;background-color:#9FDDBA">`locale = locale(...)`</span> in
your parsing function.

```{r}
parse_double("1.23")
parse_double("1,23", locale = locale(decimal_mark = ","))
```

```{r}
parse_number("$100")
parse_number("20%")
parse_number("It cost $123.45")
```

```{r}
# Used in America
parse_number("$123,456,789")
# Used in many parts of Europe
parse_number("123.456.789", locale = locale(grouping_mark = "."))
# Used in Switzerland
parse_number("123'456'789", locale = locale(grouping_mark = "'"))
```

```{r}
charToRaw("Hadley")
(x1 <- "El Ni\xf1o was particularly bad this year")
(x2 <- "\x82\xb1\x82\xf1\x82\xc9\x82\xbf\x82\xcd")
```

To fix the problem you need to specify the encoding in `parse_character()`:

```{r}
parse_character(x1, locale = locale(encoding = "Latin1"))
parse_character(x2, locale = locale(encoding = "Shift-JIS"))
```

You can try the guess_encoding() to help you out:

```{r}
guess_encoding(charToRaw(x1))
guess_encoding(charToRaw(x2))
```

```{r}
fruit <- c("apple", "banana")
parse_factor(c("apple", "banana", "bananana"), levels = fruit)
```

```{r}
parse_datetime("2010-10-01T2010")
# If time is omitted, it will be set to midnight
parse_datetime("20101010")
parse_date("2010-10-01")
library(hms)
parse_time("01:10 am")
parse_time("20:10:01")
```

```{r}
parse_date("01/02/15", "%m/%d/%y")
parse_date("01/02/15", "%d/%m/%y")
parse_date("01/02/15", "%y/%m/%d")
parse_date("1 janvier 2015", "%d %B %Y", locale = locale("fr"))
```


### Exercises

1.  What are the most important arguments to `locale()`? 

    - The date_names, for example above we specified "fr" for French 
    date names in order to parse the date.
    - decimal_mark: Due to differences in locale, you should set this
    if your decimal numbers are separated using something other than __.__.
    - grouping_mark: The default is "," since that is what is used in the US,
    but if your grouping_mark is different, you should set this argument
    for your analysis.
    - tz: The default tz is UTC, but you may want to change it to your 
    [timezone](https://en.wikipedia.org/wiki/List_of_tz_database_time_zones). 

1.  What happens if you try and set `decimal_mark` and `grouping_mark` 
    to the same character? What happens to the default value of 
    `grouping_mark` when you set `decimal_mark` to ","? What happens
    to the default value of `decimal_mark` when you set the `grouping_mark`
    to "."?
        
    ```{r, eval = FALSE}
    # decimal_mark` and `grouping_mark` to the same character
    parse_double("123,456,78", locale = locale(decimal_mark = ",",
                                            grouping_mark = ","))
    # Error: `decimal_mark` and `grouping_mark` must be different
    
    parse_number("123,456,78", locale = locale(decimal_mark = ",",
                                            grouping_mark = ","))
    # Error: `decimal_mark` and `grouping_mark` must be different
    ```
    
    ```{r}
    # What happens to the default value of `grouping_mark` 
    # when you set `decimal_mark` to ","?
    parse_number("123 456,78", locale = locale(decimal_mark = ","))
    # no grouping preserved, whitespace is not considered as group
    # so we get an incorrect parsing
    # I would gather that since we overrode decimal_mark to be
    # equal to the grouping_mark default, this removes the
    # default, and hence has to be supplied for correct parsing
    # if you also have a specific grouping character present
    
    parse_number("123 456,78", locale = locale(decimal_mark = ",",
                                               grouping_mark = " "))
    # both specified, number parsed correctly
    
    parse_number("123.456,78", locale = locale(decimal_mark = ","))
    # even though no grouping_mark specified, parse_number 
    # handles the . grouping mark well
    
    # preserve the decimals
    print(parse_number("123.456,78", 
                       locale = locale(decimal_mark = ",")),digits=10)
    
    ```
    
    Turns out the code sets the decimal_mark if it was not set, 
    or vice versa. From readr code file
    [locale.R](https://github.com/tidyverse/readr/blob/master/R/locale.R)
    
    ```{r, eval=FALSE}
      if (missing(grouping_mark) && !missing(decimal_mark)) {
        grouping_mark <- if (decimal_mark == ".") "," else "."
      } else if (missing(decimal_mark) && !missing(grouping_mark)) {
        decimal_mark <- if (grouping_mark == ".") "," else "."
      }
    ```
    
    
    ```{r}
    # What happens to the default value of `grouping_mark` 
    # when you set `decimal_mark` to ","?
    parse_double("123456,78", locale = locale(decimal_mark = ","))
    
    parse_double("123.456,78", locale = locale(decimal_mark = ","))
    
    parse_double("123.456,78", locale = locale(decimal_mark = ",",
                                               grouping_mark = "."))
    
    parse_double("123 456,78", locale = locale(decimal_mark = ",",
                                               grouping_mark = " "))
    ```
    
    Hhmm okay, so it seems like `parse_double()` is a bit more strict,
    and does not seem to like it even if we override the 
    `locale()`. This [Stack Overflow post](https://stackoverflow.com/questions/47487424/parsing-double-grouped-number-with-readrparse-double) confirms what we see here, so
    too does [this post](https://github.com/tidyverse/readr/issues/827) and [this one](https://github.com/tidyverse/readr/issues/556). The only
    perplexing thing is that when I do set the __grouping_mark__ in locale()
    why is this not considered? Because `parse_double()` also has a default 
    locale which may be overriden by locale()? `r emo::ji("confused")`
    
    ```{r}
    # What happens to the default value of `decimal_mark` 
    # when you set the `grouping_mark` to "."
    parse_number("5.123.456,78", locale = locale(grouping_mark = "."))
    # As above shows the decimal character set to , in code

    parse_number("5.123.456,78", locale = locale(decimal_mark = ",",
                                               grouping_mark = "."))
    
    problems(parse_double("5.123.456,78", 
                          locale = locale(decimal_mark = ",",
                                          grouping_mark = ".")))
    ```

    

1.  I didn't discuss the `date_format` and `time_format` options to
    `locale()`. What do they do? Construct an example that shows when 
    they might be useful.
    
```{r}

```
    

1.  If you live outside the US, create a new locale object that encapsulates 
    the settings for the types of file you read most commonly.
    
1.  What's the difference between `read_csv()` and `read_csv2()`?
    
1.  What are the most common encodings used in Europe? What are the
    most common encodings used in Asia? Do some googling to find out.

1.  Generate the correct format string to parse each of the following 
    dates and times:
    
    ```{r}
    d1 <- "January 1, 2010"
    d2 <- "2015-Mar-07"
    d3 <- "06-Jun-2017"
    d4 <- c("August 19 (2015)", "July 1 (2015)")
    d5 <- "12/30/14" # Dec 30, 2014
    t1 <- "1705"
    t2 <- "11:15:10.12 PM"
    ```

