---
title: "Chapter 9 - Tidy Data with tidyr"
author: "Vebash Naidoo"
date: "23/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r}
options(scipen=10000)
library(tidyverse)
library(flair)
library(emo)
library(lubridate)
library(magrittr)
library(tidyquant)
theme_set(theme_tq())
```

## What's __tidy__ data anyway?

1.  Each variable must have its own column.
1.  Each observation must have its own row.
1.  Each value must have its own cell.

In all the examples _tidyr::table_ to _tidyr::table4b_, only `tidyr::table1` is tidy.

```{r}
(
  # practising the read_csv function to create table1
  # just note however that table1 is in tidyr ;)
  # tidyr::table1 etc. 
  # In all honesty, I only figured this out after "practising" :P
  table1 <- read_csv("country, year, cases, population
                     Afghanistan, 1999, 745, 19987071
                     Afghanistan, 2000, 2666, 20595360
                     Brazil, 1999, 37737, 172006362
                     Brazil, 2000, 80488, 174504898
                     China, 1999, 212258, 1272915272
                     China, 2000, 213766, 1280428583")
)
```

### Working with tidy data

```{r}
table1 %>% 
  mutate(rate = cases/population * 10000)

table1 %>% 
  count(year, wt=cases) # same as group_by and sum

table1 %>% 
  group_by(year) %>% 
  summarise(sum(cases))

ggplot(table1, aes(year, cases)) +
  geom_line(aes(group = country), colour = "grey50") +
  geom_point(aes(colour = country)) +
  scale_colour_tq()

```

### Exercises

1.  Using prose, describe how the variables and observations are organised in
    each of the sample tables.
        
    ```{r}
    tidyr::table1
    ```
    
    - Each column is a variable `r emo::ji("check")`
    - Each observation is in a row `r emo::ji("check")`
    - Each value is in a cell `r emo::ji("check")` 
    - The table describes the number of cases, and the population
    count (each in its own column) for each country and year combination. 
    
    <br>
    
    ```{r}
    tidyr::table2
    ```
    
    - Each column is a variable `r emo::ji("x")`
    - Each observation is in a row `r emo::ji("check")`
    - Each value is in a cell `r emo::ji("check")` 
    - Each row contains either the number of cases, or the population
    count for each country and year combination. 
    
    <br>
    
    ```{r}
    tidyr::table3
    ```
    
    - Each column is a variable `r emo::ji("x")`
    - Each observation is in a row `r emo::ji("check")`
    - Each value is in a cell `r emo::ji("x")` 
    - Each row considers the country and year combination, and
    shows the number of cases and the population count (separated by a /) in
    one variable named `rate`.
    
    <br>
        
    ```{r}
    tidyr::table4a
    ```
    
    - Each column is a variable `r emo::ji("check")`
    - Each observation is in a row `r emo::ji("x")`
    - Each value is in a cell `r emo::ji("check")` 
    
    <br>
    
    ```{r}
    tidyr::table4b
    ```
    
    - Each column is a variable `r emo::ji("check")`
    - Each observation is in a row `r emo::ji("x")`
    - Each value is in a cell `r emo::ji("check")`  
    - Each table houses either the number of cases, or the
    population count for each country, in separated 
    columns for each year.
    
    <br> 
    
    ```{r}
    tidyr::table5
    ```
    
    - Each column is a variable `r emo::ji("x")`
    - Each observation is in a row `r emo::ji("check")`
    - Each value is in a cell `r emo::ji("x")` 
    - The table considers each year separated into century and year
    for each country and then similar to table3 it combines the
    cases and population count in one variable `rate` (separated within
    the column by /) 
    
    <br>

1.  Compute the `rate` for `table2`, and `table4a` + `table4b`. 
    You will need to perform four operations:

    1.  Extract the number of TB cases per country per year.
    1.  Extract the matching population per country per year.
    1.  Divide cases by population, and multiply by 10000.
    1.  Store back in the appropriate place.
    
    Which representation is easiest to work with? Which is hardest? Why?
    
    ```{r}
    (
      tbl1 <- tidyr::table2 %>% 
      filter(type == "cases") %>% 
      group_by(country, year) %>% 
      mutate(cases = count) %>% 
      ungroup() %>% 
      select(country, year, cases) %>% 
      arrange(country, year)
    )
    
    (
      tbl2 <- tidyr::table2 %>% 
      filter(type == "population") %>% 
      group_by(country, year) %>% 
      mutate(population = count) %>% 
      ungroup() %>% 
      select(country_temp = country, 
             year_temp = year, 
             population) %>% 
      arrange(country_temp, year_temp)
    )
    
    (
    tbl3 <- tbl1 %>% 
      bind_cols(tbl2) %>% 
      select(c(1:3,6)) %>% 
      mutate(rate = (cases / population) * 10000) %>% 
      arrange(country, year) %>% 
      select(country, year, rate) %>% 
      mutate(type = "rate",
             count = rate) %>% 
      select(c(1,2,4,5))
    )
    
    tidyr::table2 %>% 
      bind_rows(tbl3) %>% 
      mutate(count = round(count, 2)) %>% 
      arrange(country, year, type) %>% 
      gt::gt()
    ```
    
    <br> <br>
    
    ```{r}
    
    (
      tbl1_cases <- tidyr::table4a %>% 
      select(country, `1999`) %>% 
      mutate(year = 1999,
             cases  = `1999`) %>% 
      select(country, year, cases)
    )
    
    (
      tbl2_cases <- tidyr::table4a %>% 
      select(country, "2000") %>% 
      mutate(year = 2000, 
             cases = `2000`) %>% 
      select(country, year, cases)
    )
    
    (
      tbl_cases <- tbl1_cases %>% 
      bind_rows(tbl2_cases) %>% 
      arrange(country, year)
    )
    
    (
      tbl1_pop <- tidyr::table4b %>% 
      select(country, `1999`) %>% 
      mutate(year = 1999,
             population  = `1999`) %>% 
      select(country, year, population)
    )
    
    (
      tbl2_pop <- tidyr::table4b %>% 
      select(country, "2000") %>% 
      mutate(year = 2000, 
             population = `2000`) %>% 
      select(country, year, population)
    )
    
    (
      tbl_pop <- tbl1_pop %>% 
      bind_rows(tbl2_pop) %>% 
      arrange(country, year)
    )
    
    (
      tbl_rate <- tbl_cases %>% 
        bind_cols(tbl_pop) %>% 
        janitor::clean_names() %>% 
        select(country = country_1,
               year = year_2,
               cases, population) %>% 
        mutate(rate = cases / population * 10000)
    )
    
    (
      tbl_1999 <- tbl_rate %>% 
        select(country, year, rate) %>% 
        filter(year == 1999) %>% 
        mutate(`1999` = rate) %>% 
        select(country, `1999`)
    )
    
    (
      tbl_2000 <- tbl_rate %>% 
        select(country, year, rate) %>% 
        filter(year == 2000) %>% 
        mutate(`2000` = rate) %>% 
        select(country_temp = country, `2000`)
    )
    
    (
      tbl_4c <- 
        tbl_1999 %>% 
        bind_cols(tbl_2000) %>% 
        select(country, `1999`, `2000`)
    )
    
    ```
    

1.  Recreate the plot showing change in cases over time using `table2`
    instead of `table1`. What do you need to do first?
    
    ```{r}
    
    tidyr::table1
    
    ggplot(table1, aes(year, cases)) +
      geom_line(aes(group = country), colour = "grey50") +
      geom_point(aes(colour = country)) +
      scale_colour_tq()
    
    tidyr::table2
    
    table2 %>% 
      filter(type == "cases") %>% 
      ggplot(aes(year, count)) +
      geom_line(aes(group = country), colour = "grey50") +
      geom_point(aes(colour = country)) +
      scale_colour_tq()
    
    ```
    
Sometimes you will have to resolve one of two common problems:

1. One variable might be spread across multiple columns.

1. One observation might be scattered across multiple rows. 

## Pivot Longer / Gather 

`pivot_longer()` makes datasets longer by increasing the number of rows and decreasing the number of columns.

```{r tidy1, include=FALSE}
table4a

table4a %>% 
  # gather(list out columns you want to gather like dplyr::select() style,
  #        key = what do you want to call the column
  #              these column names go into,
  #        value = the values of the columns will go here)
  gather(`1999`, `2000`, 
         key = "year",
         value = "cases" )

(tidy_4a <- table4a %>% 
  # cols = list the columns you want to pivot
  # names_to = what will you call the new column these
  #            column names go into
  # values_to = the values in the columns will go here
  pivot_longer(cols = c(`1999`, `2000`),
               names_to = "year",
               values_to = "cases"))
```

```{r, echo=FALSE}
decorate("tidy1") %>% 
  flair("gather", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("pivot_longer", background = "#9FDDBA", 
        color = "#008080")  
```

```{r tidy2, include=FALSE}
table4b

table4b %>% 
  gather(`1999`, `2000`, 
         key = "year",
         value = "population")

(tidy_4b <- table4b %>% 
  pivot_longer(cols = c(`1999`, `2000`),
               names_to = "year",
               values_to = "population"))

left_join(tidy_4a, tidy_4b) %>% 
  arrange(country, year)
```

```{r, echo=FALSE}
decorate("tidy2") %>% 
  flair("gather", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("pivot_longer", background = "#9FDDBA", 
        color = "#008080")
  
```

## Pivot Wider / Spread

`pivot_wider()` is the opposite of `pivot_longer()`. You use it when an observation is scattered across multiple rows.

```{r tidy3, include=FALSE}
table2

table2 %>% 
  # key = column with the variable name, here `type`
  spread(key = type, 
  # value = column with the value that will be assigned
  # to new columns
         value = count)

table2 %>% 
  pivot_wider(names_from = type,
              values_from = count)
```

```{r, echo=FALSE}
decorate("tidy3") %>% 
  flair("spread", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("pivot_wider", background = "#9FDDBA", 
        color = "#008080")
  
```

### Exercises

1.  Why are `pivot_longer()` and `pivot_wider()` not perfectly symmetrical?  
    Carefully consider the following example:
    
    ```{r}
    (stocks <- tibble(
      year   = c(2015, 2015, 2016, 2016),
      half  = c(   1,    2,     1,    2),
      return = c(1.88, 0.59, 0.92, 0.17)
    ))
    stocks %>% 
      pivot_wider(names_from = year, values_from = return) %>% 
      pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return")
    ```
    
    (Hint: look at the variable types and think about column _names_.)
    
    `pivot_longer()` has a `names_ptypes` argument, e.g. 
    `names_ptypes = list(year = double())`. What does it do?
    
    ```{r}
    # vignette("pivot")
    stocks %>% 
        pivot_wider(names_from = year, values_from = return)
    ```
    
    Let's have a look at the first part - here we take the year
    and make it a variable. That means that `2015` and `2016` become
    variables (new columns) in our new tibble, and the return
    gets pulled into the appropriate column (`2015`/`2016`) against the
    appropriate `half`. By nature of this move we changed year which was a 
    double into two new column names which are `2015` and `2016` and hence
    "character".
    
    ```{r}
    (stocks_ <- stocks %>% 
      pivot_wider(names_from = year, values_from = return) %>% 
      pivot_longer(`2015`:`2016`, names_to = "year", values_to = "return"))
    colnames(stocks)
    colnames(stocks_)
    ```
    
    So following on that we take these new columns and then collapse
    them into a column `year` again. But now we have changed the type
    given we made them columns in the `pivot_wider()` step. So they keep
    their "character" nature when they are made longer again. Final result
    is `year` started off double (when we created it) but ends up character
    (after the pivot_wider and pivot_longer steps).
    
    Th columns also get rearranged since the pivot_wider spreads the `year`
    column into `2015` and `2016` which come after `half` in that
    initial step. 
    When we subsequently pivot_longer `half` remains as the first 
    column, followed by the `names_to = ` column (year in this case), and 
    finally the `values_to = ` column (return in this case).
    
    __Q__: `pivot_longer()` has a `names_ptypes` argument, e.g. 
    `names_ptypes = list(year = double())`. What does it do?
    
    Okay so upon reading the help page and the [info](https://rdrr.io/github/tidyverse/tidyr/man/pivot_longer.html)
    I expected that this function would convert my `character`
    column year created after the pivot_wider() step into a double, 
    but instead it throws an error. `r emo::ji("confused")`
    
    ```{r, error = TRUE}
    stocks %>% 
      pivot_wider(names_from = year, values_from = return) %>% 
      pivot_longer(`2015`:`2016`, 
                   names_to = "year", 
                   names_ptypes = list(year = double()),
                   values_to = "return"
                   )
    ```

    We use this to __confirm__ that the columns we create are
    of the type / class we expect - so here it provides a <span style="color: #008080;background-color:#9FDDBA">__check__</span> it seems
    `r emo::ji("shrug")`. <br>
    
    To transform the column from character to double you would need
    to use the <span style="color: #008080;background-color:#9FDDBA">`names_transform`</span>.
    
        
    ```{r}
    (stocks_ptypes <- stocks %>% 
      pivot_wider(names_from = year, values_from = return) %>% 
      pivot_longer(`2015`:`2016`, 
                   names_to = "year", 
                   names_transform = list(year = as.double),
                   values_to = "return",
                   # is the value column of the type expected
                   values_ptypes = list(return = double())
                   ))
    ```
    
    Strangely though I would expect that if I transform a column
    from x to y (using `names_transform`), and then use `names_ptypes` 
    to check if my name column is indeed now of type y that would be fine? 
    It still throws an error, so my thinking is flawed here.
    
    ```{r, error = TRUE}
    stocks %>% 
      pivot_wider(names_from = year, values_from = return) %>% 
      pivot_longer(`2015`:`2016`, 
                   names_to = "year", 
                   names_transform = list(year = as.double),
                   names_ptypes = list(year = double()),
                   values_to = "return",
                   # is the value column of the type expected
                   values_ptypes = list(return = double())
                   )
    ```
    


1.  Why does this code fail?

    ```{r, error = TRUE}
    table4a %>% 
      pivot_longer(c(1999, 2000), names_to = "year", values_to = "cases")
      # Error: Can't subset columns that don't exist.
      # x Locations 1999 and 2000 don't exist.
      # i There are only 3 columns.
    ```
    
    ```{r}
    # fixing it
    table4a %>% 
      pivot_longer(c("1999", `2000`), names_to = "year", values_to = "cases")
    ```
    
    The `1999` and `2000` are non-syntactically named columns. These 
    have to be surrounded by backticks (\`\`) or quotations `""`. Here tidyr
    is trying to read columns numbered 1999, and 2000 which don't
    exist.

1.  What would happen if you widen this table? Why? How could you add a 
    new column to uniquely identify each value?

    ```{r}
    people <- tribble(
      ~name,             ~names,  ~values,
      #-----------------|--------|------
      "Phillip Woods",   "age",       45,
      "Phillip Woods",   "height",   186,
      "Phillip Woods",   "age",       50,
      "Jessica Cordero", "age",       37,
      "Jessica Cordero", "height",   156
    )
    ```
    
    You get a warning and it has a list for each variable `age` 
    and `height` since _Philips Woods_ has two ages which are different.
        
    ```{r}
    people %>% 
      pivot_wider(names_from = names,
                  values_from = "values")
    ```

    ```{r}
    people2 <- tribble(
      ~name,             ~names,  ~values,
      #-----------------|--------|------
      "Phillip Woods",   "age",       45,
      "Phillip Woods",   "height",   186,
      "Phillip Woods",   "age2",      50, # second age gets diff col name
      "Jessica Cordero", "age",       37,
      "Jessica Cordero", "height",   156
    )
    ```
    
    ```{r}
    people2 %>% 
      pivot_wider(names_from = names,
                  values_from = "values")
    ```

1.  Tidy the simple tibble below. Do you need to make it wider or longer?
    What are the variables?

    ```{r}
    (preg <- tribble(
      ~pregnant, ~male, ~female,
      "yes",     NA,    10,
      "no",      20,    12
    ))
    ```
    
    We need to make it longer. The variable's are `pregnant` (yes or no), 
    and the number of male(s)/female(s) in each outcome of `pregnant`.
    
    ```{r}
    preg %>% 
      pivot_longer(c('male', 'female'), 
                   names_to = 'sex',
                   values_to = 'count')
    ```
    
## Separating and Uniting

### Separate

`separate()` pulls apart one column into multiple columns, by splitting wherever a separator character appears, and by default, `separate()` will split values wherever it sees a non-alphanumeric character (i.e. a character that isn't a number or letter e.g. `,`, `;`, etc.).

In `tidyr::table3` we have a column `rate` that has a combination of the
cases and population in one column separated by a _/_.

```{r}
table3
```

#### No __sep__ specified

```{r sep1, include=FALSE}
table3 %>% 
  # separate without specifying the separator
  separate(rate, into = c("cases", "population"))
```

```{r, echo=FALSE}
decorate("sep1") %>% 
  flair("separate", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("into =", background = "#9FDDBA", 
        color = "#008080")
```

#### __sep__ specified as single value

<span style="color: #008080;background-color:#9FDDBA">`sep`</span> is
a regular expression, so more complicated sequences than "/" maybe supplied,
e.g. `sep = '[0-9]'` will separate on any digit encountered.

```{r sep2, include=FALSE}
table3 %>% 
  # separate with specifying the separator
  separate(rate, into = c("cases", "population"),
           sep = '/')

as_tibble("This0is1a2test.with3a4deliberate5mistake.") %>%
  separate(value, into = c(glue::glue("col_{c(1:7)}")),
         sep = '[0-9]')
```

```{r, echo=FALSE}
decorate("sep2") %>% 
  flair("sep = '/'", background = "#9FDDBA", 
        color = "#008080") 
```


#### __sep__ specified as vector of integers / position

You can also pass a vector of integers to `sep`. `separate()` will interpret the integers as positions to split at.

- Positive values start at 1 and start counting from the left of the string
- Negative values start at -1 and start counting from the right of the string


```{r sep4, include=FALSE}
table3 %>% 
  # separate with specifying the separator
  separate(year, into = c("century", "year"),
           # here we use a single value
           # use c(2, 5) to split in multiple places
           sep = 2,
           convert = TRUE)
```

```{r, echo=FALSE}
decorate("sep4") %>% 
  flair("sep = 2", background = "#9FDDBA", 
        color = "#008080") 
```

#### convert = TRUE

`separate()` retains the character type of the original field.
Use <span style="color: #008080;background-color:#9FDDBA">`convert = TRUE`</span> to get `separate()` to make a guess at a better type
for the resulting columns. 

```{r sep3, include=FALSE}
table3 %>% 
  # separate with specifying the separator
  separate(rate, into = c("cases", "population"),
           sep = '/',
           convert = TRUE)
```

```{r, echo=FALSE}
decorate("sep3") %>% 
  flair("convert = TRUE", background = "#9FDDBA", 
        color = "#008080") 
```

### Unite

`unite()` is the inverse of `separate()`: it combines multiple columns into a single column.

The year column was deliberately made untidy by splitting it into century and year in `tidyr::table5`. We can concat these using <span style="color: #008080;background-color:#9FDDBA">`unite`</span>.

```{r unite1, include=FALSE}
table5 %>% 
  unite(new, century, year)
```

```{r, echo=FALSE}
decorate("unite1") %>% 
  flair("unite", background = "#9FDDBA", 
        color = "#008080")
```

If you look at the help page of unite (?unite) you will see that
the default separator is ___ (`sep = "_"`). To override this behaviour
we need to specify the `sep` argument.


```{r unite2, include=FALSE}
table5 %>% 
  unite(new, century, year, sep = '')
```

```{r, echo=FALSE}
decorate("unite2") %>% 
  flair("sep = ''", background = "#9FDDBA", 
        color = "#008080")  
```

### Exercises

1.  What do the `extra` and `fill` arguments do in `separate()`? 
    Experiment with the various options for the following two toy datasets.
    
    ```{r, eval = FALSE}
    tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
      separate(x, c("one", "two", "three"))
    
    tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
      separate(x, c("one", "two", "three"))
    ```
    
    - The default for _extra_ is "warn".
    - `extra` warns if during the splitting more columns are found than
    those specified in _into_. For example I may have 
    `into = c("year_of_birth", "month_of_birth")` but upon splitting on `-`
    three columns are found. If I don't specify `extra` then _separate_
    function will warn me there are more columns found, but these will be
    missing from my result.
    - `extra` can be _warn_ (default), _drop_ (drop any extra values
    without a warning) or _merge_ (only splits at most length(into) times).
    
    ```{r ex1_1, include=FALSE, warning = TRUE}
    tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
      separate(x, c("one", "two", "three"))
    tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
      separate(x, c("one", "two", "three"),
               extra = 'drop')
    tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
      separate(x, c("one", "two", "three"),
               extra = 'merge')
    ```

    ```{r, echo=FALSE}
    decorate("ex1_1") %>% 
      flair("extra = 'drop'", background = "#9FDDBA", 
            color = "#008080") %>% 
      flair("extra = 'merge'", background = "#9FDDBA", 
            color = "#008080")
    ``` 
    
    - The default for _fill_ is "warn".
    - `fill` warns if during the splitting less columns are found than
    those specified in _into_. <br>
    For example I may have 
    `into = c("year_of_birth", "month_of_birth", "day_of_birth", "sex")` 
    but upon splitting on `-` only three columns are found. 
    If I don't specify `fill` then the _separate_
    function will warn me there are less columns than expected 
    and these will be filled with NA.
    - `fill` can be:
    
      * _warn_ (default)
      * _right_ (fill with missing values on the right) - i.e. if there are
      not enough values for the columns in the tibble the right hand side
      columns must be filled with NA. 
      You'd use this if you expect that more of the end columns are
      missing.
      * _left_ (fill with missing values on the left) - i.e. if there are
      not enough values for the columns in the tibble the left hand side
      columns must be filled with NA. 
      You'd use this if you expect that more of the beginning columns
      you're making are likely to be missing.
      
    ```{r ex1_2, include=FALSE, warning = TRUE}
    tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
      separate(x, c("one", "two", "three"))
    tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
      separate(x, c("one", "two", "three"),
               fill = 'right')    
    tibble(x = c("a,b,c", "d,e", "f,g,i")) %>% 
      separate(x, c("one", "two", "three"),
               fill = 'left')    
    ```

    ```{r, echo=FALSE}
    decorate("ex1_2") %>% 
      flair("fill = 'right'", background = "#9FDDBA", 
            color = "#008080") %>% 
      flair("fill = 'left'", background = "#9FDDBA", 
            color = "#008080")
    ``` 
    
1.  Both `unite()` and `separate()` have a `remove` argument. What does it
    do? Why would you set it to `FALSE`?
    
    - `remove` is by default TRUE.
    - It means remove the column(s) from which you are 
    separating / uniting.
    - You would set it to _FALSE_ if you'd like to perform the action
    (separate / unite) but __also__ keep the original column(s).
    
    ```{r ex2, include=FALSE, warning = TRUE}
    tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
      separate(x, c("one", "two", "three"))
    tibble(x = c("a,b,c", "d,e,f,g", "h,i,j")) %>% 
      separate(x, c("one", "two", "three"),
               remove = FALSE)
    ```

    ```{r, echo=FALSE}
    decorate("ex2") %>% 
      flair("remove = FALSE", background = "#9FDDBA", 
            color = "#008080")
    ```     
    
    Notice that initially `x` does not exist. In the second call we
    added the <span style="color: #008080;background-color:#9FDDBA">remove = FALSE</span> and `x` is kept.
    
1.  Compare and contrast `separate()` and `extract()`.  Why are there
    three variations of separation (by position, by separator, and with
    groups), but only one unite?
    
    - separate has by position, by separator, and with groups because it
    takes a string and splits it into multiple parts; unite on the other
    hand combines separate columns and there is only one way to do
    it - specify the columns you want to join, in the order
    they should be joined, and with the separator that is appropriate.
    
    - separate does not throw away any part of the string unless it
    "falls over the edge" so to say. I.e. if we have extra string left
    over after the splitting, and it has run out of columns to place it
    in. You may kinda simulate this by saying "Hey, keep only the
    second column" as this help example shows. While __sep__ is a 
    regex it tells you how to split the string.
    
      ```{r}
      # If you want to split by any non-alphanumeric value (the default):
      df <- data.frame(x = c(NA, "a.b", "a.d", "b.c"))
      df %>% separate(x, c("A", "B"))
      
      # If you just want the second variable:
      df %>% separate(x, c(NA, "B"))
      
      ```
      
    - extract will give you back only the "groups" you are interested
    in, and "throws" away the rest, and maybe better to use when
    you have complex patterns you want to extract. `r emo::ji("shrug")`
    
    ```{r}
    df <- tibble(x = c(NA, "a-b", "a-d", "b-c", "d-e"))
    
    df %>% tidyr::extract(x,
                         # make 2 columns
                         c("A", "B"), 
                         # here's the groups I want - surrounded by ()
                         "([a-d]+)-([a-d]+)")
    
    df %>% 
      tidyr::extract(col = x,
                    into = c("A"),
                    regex = "([a-d]+)")
    
    df <- tibble(x = c(NA, "a-b-c", "a-d-e", "b-c-d", "d-e-f"))
    
    df %>% tidyr::extract(x,
                         # make 2 columns
                         c("A", "C"), 
                         # here's the groups I want - surrounded by ()
                         # only first, and last
                         "([a-z]+)-[a-z]-([a-z]+)")
    
    
    df %>% tidyr::separate(x,
                         # You can also do it using separate
                         # make 2 columns
                         c("A", "C"), 
                         # here's the groups I want - surrounded by ()
                         # only first, and last
                         "-[a-z]-")
    
    ```

### Missing values    

A value can be missing in one of two possible ways:

* __Explicitly__, i.e. flagged with `NA`.
* __Implicitly__, i.e. simply not present in the data.

For example here we have explicit _NA_ as well as _2016_ is missing
quarter 1.

```{r miss, include=FALSE}
(stocks <- tibble(
  year   = c(2015, 2015, 2015, 2015, 2016, 2016, 2016),
  qtr    = c(   1,    2,    3,    4,    2,    3,    4),
  return = c(1.88, 0.59, 0.35,   NA, 0.92, 0.17, 2.66)
))
```

```{r, echo=FALSE}
decorate("miss") %>% 
  flair("NA", background = "#9FDDBA", 
        color = "#008080") %>% 
  flair("2016", background = "#9FDDBA", 
        color = "#008080")
```

Representing the dataset can make the missing values more clear.

1. Show implicit missing values.
    
    ```{r}
    stocks %>% 
      pivot_wider(names_from = year, values_from = return)
    ```
    
1. Turn explicit missing values, into implicit missing values.

    ```{r}
    stocks %>% 
      pivot_wider(names_from = year, values_from = return) %>% 
      pivot_longer(
        cols = c(`2015`, `2016`), 
        names_to = "year", 
        values_to = "return", 
        values_drop_na = TRUE
      )
    ```

1. Use <span style="color: #008080;background-color:#9FDDBA">`complete()`</span> : make missing values explicit, by looking
at a set of columns and finding all combinations. The gaps are filled
with NA explicitly.

    ```{r}
    stocks %>% 
      complete(year, qtr)
    ```

1. Sometimes when working with data the missing are an indication
that the previous rows value must be filled down. 
    
    ```{r}
    (treatment <- tribble(
      ~ person,           ~ treatment, ~response,
      "Derrick Whitmore", 1,           7,
      NA,                 2,           10,
      NA,                 3,           9,
      "Katherine Burke",  1,           4
    ))
    ```
    
    Using <span style="color: #008080;background-color:#9FDDBA">`fill()`</span> allows you to fill 
    a columns values with the most recent non-missing value.
    
    ```{r}
    treatment %>% 
      fill(person)
    ```    

### Exercises

1.  Compare and contrast the `fill` arguments to `pivot_wider()` and `complete()`. 

    - Both fill missing values with something other than NA.
    - In pivot_wider() it is called `values_fill`, and the spec
    is either a `scalar` or `named list`.
    - In complete() it is called fill, and the spec
    is a `named list`.

    ```{r}
    (df <- tibble(
         group = c(1:2, NA_real_, 1),
         item_id = c(1:2, 1, 2),
         item_name = c("a", "NA", "b", "b"),
         value1 = 1:4,
         value2 = 4:7
     ))
    
    df %>% 
      pivot_wider(names_from = item_name, 
                  values_from = value1,
                  # the values are filled with 0 if missing
                  values_fill = 0)
    
    df %>% 
      complete(group, item_id,
               # explicitly list cols you want filled
               # with alternate to NA
               fill = list(group = 0,
               item_name = "0"))
    ```

1.  What does the direction argument to `fill()` do?    

    The argument is `.direction` and tells you in which 
    direction the fill() must occur, and can be one of 4
    different values.
    
      > .direction = c("down", "up", "downup", "updown")

    ```{r}
    # Value (n_squirrels) is missing above and below within a group
    (squirrels <- tibble::tribble(
      ~group,    ~name,     ~role,     ~n_squirrels,
      1,      "Sam",    "Observer",   NA,
      1,     "Mara", "Scorekeeper",    8,
      1,    "Jesse",    "Observer",   NA,
      1,      "Tom",    "Observer",   NA,
      2,     "Mike",    "Observer",   NA,
      2,  "Rachael",    "Observer",   NA,
      2,  "Sydekea", "Scorekeeper",   14,
      2, "Gabriela",    "Observer",   NA,
      3,  "Derrick",    "Observer",   NA,
      3,     "Kara", "Scorekeeper",    9,
      3,    "Emily",    "Observer",   NA,
      3, "Danielle",    "Observer",   NA
    ))
    
    # The values are inconsistently missing by position within the group
    # Use .direction = "downup" to fill missing values in both directions
    squirrels %>%
      dplyr::group_by(group) %>%
      fill(n_squirrels, .direction = "downup") %>%
      dplyr::ungroup()
    
    # contrast with "updown"
    squirrels %>%
      dplyr::group_by(group) %>%
      # should be the same in this e.g.
      fill(n_squirrels, .direction = "updown") %>%
      dplyr::ungroup()
    
    # contrast with "down"
    squirrels %>%
      dplyr::group_by(group) %>%
      fill(n_squirrels, .direction = "down") %>%
      dplyr::ungroup()
    
    # contrast with "up"
    squirrels %>%
      dplyr::group_by(group) %>%
      fill(n_squirrels, .direction = "up") %>%
      dplyr::ungroup()
    ```
    
### Case Study

```{r}
who
```

```{r}
(
  who1 <- who %>% 
          # change the wide data to long data
          pivot_longer(
            cols = new_sp_m014:newrel_f65, 
            names_to = "key", 
            values_to = "cases", 
            values_drop_na = TRUE
          )
)
```

```{r}
who1 %>% 
  # how many keys do we have?
  count(key, sort = TRUE) %>% 
  DT::datatable()
```

To parse the columns, refer to Chaper 9 ([Chapter 12 online](https://r4ds.had.co.nz/tidy-data.html#case-study)).

```{r}
(who2 <- who1 %>% 
  # Look in key. Do you see any strings that have "newrel"?
  # If you do replace these with new_rel for consistent
  # key values for easier parsing
  mutate(key = stringr::str_replace(key,
                                    "newrel",
                                    "new_rel")))
```

```{r}
(
  who3 <- who2 %>% 
    separate(key,
             c("new", "type", "sexage"),
             sep = "_")
)
```

Ok, from the chapter text it seems we only have __new__ cases, so let us
verify that.

```{r}
who3 %>% 
  count(new)
```

This is indeed the case! The count == the dataset count so we can drop
this column

```{r}
(
  who4 <- who3 %>% 
    select(-new, -iso2, iso3)
)
```

Let's separate the `sexage` column.

```{r}
(
  who5 <- who4 %>% 
    separate(sexage, c("sex", "age"),
             sep = 1)
)
```

__Pipeline version__

```{r}
who %>% 
  pivot_longer(cols = new_sp_m014:newrel_f65,
               names_to = "key",
               values_to = "cases",
               values_drop_na = TRUE) %>% 
  mutate(key = stringr::str_replace(key, "newrel", "new_rel")) %>% 
  separate(key, c("new", "type", "sexage"),
                  sep = "_") %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"),
           # first pos = sex, rest = age group
           sep = 1)
```

### Exercises

1.  In this case study I set `values_drop_na = TRUE` just to make it easier to
    check that we had the correct values. Is this reasonable? Think about
    how missing values are represented in this dataset. Are there implicit
    missing values? What's the difference between an `NA` and zero? 
    
    If we consider the missing values in this data set, it does
    seem reasonable.
    
    ```{r}
    who %>% 
      select(new_sp_m014:newrel_f65) %>% 
      skimr::skim()

    who %>% 
      select(new_sp_m014) %>% 
      filter(new_sp_m014 == 0 |
             is.na(new_sp_m014)) %>% 
      skimr::skim()   
    
    who %>% 
      select(new_sp_m014) %>% 
      filter(new_sp_m014 == 0) %>% 
      skimr::skim()   
    
    who %>% 
      select(new_sp_m014) %>% 
      filter(is.na(new_sp_m014)) %>% 
      skimr::skim()   
    
    ```
    
    There may be implicit values if a year is missing for a country.
    
    __0__ means I received a measure for that variable and it is no cases.
    __NA__ could mean I received no information for that variable so
    cases may or may not exist.

1.  What happens if you neglect the `mutate()` step?
    (`mutate(names_from = stringr::str_replace(key, "newrel", "new_rel"))`)
    
    For some of the steps we will have misaligned data when we
    separate these later in the pipeline. 
    
    Let's take for example `newrel_m3544`
    
    - in the separate(key, ...) step _new = newrel_, _type = m3544_
    - this will complicate the dropping of "new"
    - it will also mean that type would be m3544.
    

1.  I claimed that `iso2` and `iso3` were redundant with `country`. 
    Confirm this claim.
    
    ```{r}
    who %>% 
      count(country, sort = TRUE)
    
    who %>% 
      count(iso2, sort = TRUE)
    
    who %>% 
      count(iso3, sort = TRUE)
    
    who %>% 
      select(country, iso2, iso3) %>% 
      skimr::skim()
    ```
    
    In the skim for character variables the min tells you the min chars in
    a observation, and max is the max chars in a variable.
    

1.  For each country, year, and sex compute the total number of cases of 
    TB. Make an informative visualisation of the data.
    
    ```{r}
    who5 %>% 
      group_by(country, year, sex) %>% 
      summarise(total_cases = sum(cases)) %>% 
      ungroup() %>% 
      arrange(country, year, sex) %>% 
      filter(country %in% c("South Africa", "Botswana",
                           "Zimbabwe", "Namibia", "Swaziland",
                           "Lesotho", "Mozambique")) %>% 
      ggplot(aes(x = year, y = total_cases, colour = sex)) +
      geom_point(alpha = 0.5) +
      facet_wrap(~ country, scales = "free_y") +
      labs(title = "Total Cases Across South Africa 
           and neighbouring countries over time",
           subtitle = "The number of cases is growing quite a bit over time",
           x = "", 
           y = "Cases Per Year") +
      scale_colour_tq()
    
    who5 %>% 
      group_by(country, year, sex) %>% 
      summarise(total_cases = sum(cases)) %>% 
      ungroup() %>% 
      arrange(country, year, sex) %>% 
      group_by(year, sex) %>% 
      summarise(cases_per_year = sum(total_cases)) %>% 
      ungroup() %>% 
      ggplot(aes(x = year, y = cases_per_year, fill = sex)) +
      geom_col(alpha = 0.5) +
      scale_fill_tq() +
      scale_y_log10() +
      labs(title = "Total Cases Across all countries over time",
           subtitle = "The number of cases is growing quite a bit over time",
           x = "", 
           y = "Cases Per Year")
    
    ```
    

















